---
title: "energy_1"
output: html_document
date: "2025-04-30"
---

Create graphs related to energy
-Relate gdp/cap to energy prduction and consumption

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}

```



```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(fmsb)
library(tibble)
library(plotly)
```


Load Data

```{r}
wfb_df <- read.csv('cleaned_wfb.csv')
head(wfb_df)
```



Clean Energy..Coal...production and convert to numeric
Convert metric tons to BTU

```{r}
head(wfb_df$Energy..Coal...production)

clean_coal_column <- function(x) {
  # Remove text like 'metric tons' and any parentheses content
  x <- gsub(" metric tons.*", "", x)  # Remove 'metric tons' and anything after it
  x <- gsub("\\(.*\\)", "", x)         # Remove anything in parentheses like '(2022 est.)'
  
  # Convert 'million' to 1e6 and 'billion' to 1e9
  x <- gsub(" million", "e6", x)      # Convert 'million' to 'e6'
  x <- gsub(" billion", "e9", x)      # Convert 'billion' to 'e9'
  x <- gsub(" trillion", "e12", x)      # Convert 'trillion' to 'e12'
  
  # Remove any commas and extra spaces
  x <- gsub(",", "", x)  # Remove commas
  
  # Convert to numeric after handling the units
  return(as.numeric(x))  # Convert the cleaned string to numeric
}

wfb_df$Energy..Coal...production = sapply(wfb_df$Energy..Coal...production, clean_coal_column)

head(wfb_df$Energy..Coal...production)

#coal in metric tonnes
#using 11,500 BTU/lb according to NC State
wfb_df$Energy..Coal...production = wfb_df$Energy..Coal...production *2204.62*11500
```



Clean Energy..Natural.gas...production and convert to numeric
Convert cubic meters to BTU


```{r}
#Clean Energy..Natural.gas...production and convert to numeric
head(wfb_df$Energy..Natural.gas...production)

clean_natural_gas_column <- function(x) {
  # Remove text like 'cubic meters' and any parentheses content
  x <- gsub(" cubic meters.*", "", x)  # Remove 'metric tons' and anything after it
  x <- gsub("\\(.*\\)", "", x)         # Remove anything in parentheses like '(2022 est.)'
  
  # Convert 'million' to 1e6 and 'billion' to 1e9
  x <- gsub(" million", "e6", x)      # Convert 'million' to 'e6'
  x <- gsub(" billion", "e9", x)      # Convert 'billion' to 'e9'
  x <- gsub(" trillion", "e12", x)      # Convert 'trillion' to 'e12'
  
  # Remove any commas and extra spaces
  x <- gsub(",", "", x)  # Remove commas
  
  # Convert to numeric after handling the units
  return(as.numeric(x))  # Convert the cleaned string to numeric
}

wfb_df$Energy..Natural.gas...production = sapply(wfb_df$Energy..Natural.gas...production, clean_natural_gas_column)

head(wfb_df$Energy..Natural.gas...production)

#natural gas in cubic meters
#34,121 BTU/m^3 according to BP
#i never cleaned this column yet
wfb_df$Energy..Natural.gas...production = wfb_df$Energy..Natural.gas...production * 34121
```




Clean Energy..Petroleum.total.petroleum.production and convert to numeric
Convert bbl/day to BTU
```{r}
#Clean Energy..Petroleum...total.petroleum.production and convert to numeric
head(wfb_df$Energy..Petroleum...total.petroleum.production)

clean_petroleum_column <- function(x) {
  # Remove text like 'metric tons' and any parentheses content
  x <- gsub(" bbl/day.*", "", x)  # Remove 'metric tons' and anything after it
  x <- gsub("\\(.*\\)", "", x)         # Remove anything in parentheses like '(2022 est.)'
  
  # Convert 'million' to 1e6 and 'billion' to 1e9
  x <- gsub(" million", "e6", x)      # Convert 'million' to 'e6'
  x <- gsub(" billion", "e9", x)      # Convert 'billion' to 'e9'
  x <- gsub(" trillion", "e12", x)      # Convert 'trillion' to 'e12'
  
  # Remove any commas and extra spaces
  x <- gsub(",", "", x)  # Remove commas
  
  # Convert to numeric after handling the units
  return(as.numeric(x))  # Convert the cleaned string to numeric
}

wfb_df$Energy..Petroleum...total.petroleum.production = sapply(wfb_df$Energy..Petroleum...total.petroleum.production, clean_petroleum_column)

head(wfb_df$Energy..Petroleum...total.petroleum.production)


#petroleum in barrels per day
#5.8e6 BTU per barrel of oil according to BP
wfb_df$Energy..Petroleum...total.petroleum.production = wfb_df$Energy..Petroleum...total.petroleum.production * 365 * 5.8e6
```

Make aggregate of production(coal+natural_gas+petroleum)
```{r}
#make aggregate of production
wfb_df <- wfb_df %>%
  mutate(total_production = rowSums(select(., Energy..Coal...production, Energy..Natural.gas...production, Energy..Petroleum...total.petroleum.production), na.rm = TRUE))
head(wfb_df$total_production)
```

Clean GDP per capita column

```{r}
#Clean and convert column: Economy..Real.GDP.per.capita
head(wfb_df$Economy..Real.GDP.per.capita)

clean_gdp_column <- function(x) {
  # Remove dollar sign and commas
  x <- gsub("^\\$", "", x)  # Remove the first dollar sign
  x <- gsub(",", "", x)     # Remove the commas
  
  # Remove everything after the first numeric value (remove text like 'est.', etc.)
  x <- gsub(" .*", "", x)   # Remove everything after the first number
  
  # Convert to numeric and return
  return(as.numeric(x))  # Convert cleaned string to numeric
}

wfb_df$Economy..Real.GDP.per.capita = sapply(wfb_df$Economy..Real.GDP.per.capita, clean_gdp_column)

head(wfb_df$Economy..Real.GDP.per.capita)
```


Clean column: Energy Consumption per capita

```{r}
#Clean and convert Energy..Energy.consumption.per.capita column
head(wfb_df$Energy..Energy.consumption.per.capita)

clean_consumption_column <- function(x) {
  # Remove text like 'metric tons' and any parentheses content
  x <- gsub(" Btu/person.*", "", x)  # Remove 'metric tons' and anything after it
  x <- gsub("\\(.*\\)", "", x)         # Remove anything in parentheses like '(2022 est.)'
  
  # Convert 'million' to 1e6 and 'billion' to 1e9
  x <- gsub(" million", "e6", x)      # Convert 'million' to 'e6'
  x <- gsub(" billion", "e9", x)      # Convert 'billion' to 'e9'
  x <- gsub(" trillion", "e12", x)      # Convert 'trillion' to 'e12'
  
  # Remove any commas and extra spaces
  x <- gsub(",", "", x)  # Remove commas
  
  # Convert to numeric after handling the units
  return(as.numeric(x))  # Convert the cleaned string to numeric
}

wfb_df$Energy..Energy.consumption.per.capita = sapply(wfb_df$Energy..Energy.consumption.per.capita, clean_consumption_column)


head(wfb_df$Energy..Energy.consumption.per.capita)
```

Clean column: People.and.Society..Population...total
```{r}
head(wfb_df$People.and.Society..Population...total)

clean_population_column <- function(x) {
  x <- gsub(",", "", x)     # Remove the commas

  # Convert to numeric and return
  return(as.numeric(x))  # Convert cleaned string to numeric
}

wfb_df$People.and.Society..Population...total = sapply(wfb_df$People.and.Society..Population...total, clean_population_column)

head(wfb_df$People.and.Society..Population...total)
```

Scale energy production to per capita. Save in column: Energy..Energy.production.per.capita

```{r}
wfb_df$Energy..Energy.production.per.capita =
(wfb_df$total_production)/(wfb_df$People.and.Society..Population...total)

head(wfb_df$Energy..Energy.production.per.capita)
```

Fix continent names
```{r}
wfb_df$Geography..Map.references[wfb_df$Geography..Map.references == "Arctic Region"] <- "Europe"
wfb_df$Geography..Map.references[wfb_df$Geography..Map.references == "AsiaEurope"] <- "Europe"
wfb_df$Geography..Map.references[wfb_df$Geography..Map.references == "Southeast Asia"] <- "Asia"
wfb_df$Geography..Map.references[wfb_df$Geography..Map.references == "Central America and the Caribbean"] <- "South America"
wfb_df$Geography..Map.references[wfb_df$Country == "France"] <- "Europe"

wfb_df$Geography..Map.references[wfb_df$Country == "World"] <- "World"
```


Make graph

```{r}
#change 'size' parameter for different modes of production
#Choose from: 'total_production', 'Energy..Energy.production.per.capita', 'Energy..Coal...production', Energy..Petroleum...total.petroleum.production", or 'Energy..Natural.gas...production' columns 
ggplot(wfb_df, aes(
  x = Economy..Real.GDP.per.capita,
  y = Energy..Energy.consumption.per.capita,
  size = total_production,
  label = Country,
  color = Geography..Map.references
)) +
  geom_point(alpha = 0.7) +
  geom_text(check_overlap = TRUE, size = 3, vjust = -1, show.legend = FALSE) +
  scale_color_brewer(palette = "Set1") +
  scale_size_continuous(range = c(2, 12)) +  # adjust bubble sizes
  theme_minimal() +
  labs(
    x = "Real GDP per Capita",
    y = "Consumption per Capita (BTU)",
    size = "Total Production (BTU)",
    title = ""
  ) +
  xlim(0, 100000) +  # Set the x-axis limit (example: 0 to 50000)
  ylim(0, 350000000)     # Set the y-axis limit (example: 0 to 5000)

```

```{r}
p <- ggplot(wfb_df, aes(
  x = Economy..Real.GDP.per.capita,
  y = Energy..Energy.consumption.per.capita,
  size = Energy..Energy.production.per.capita,
  text = Country,
  color = Geography..Map.references
)) +
  geom_point(alpha = 0.7) +
  scale_color_brewer(palette = "Set1") +
  scale_size_continuous(range = c(2, 12)) +  # adjust bubble sizes
  theme_minimal() +
  labs(
    x = "Real GDP per Capita",
    y = "Consumption per Capita (BTU)",
    size = "Production per Capita (BTU)",
    title = ""
  ) +
  xlim(0, 100000) +  # Set the x-axis limit (example: 0 to 50000)
  ylim(0, 350000000)     # Set the y-axis limit (example: 0 to 5000)

ggplotly(p, tooltip = "text")
```


Grouped bar charts showing what type of energy is being produced by each continent
```{r}
long_df <- wfb_df %>%
  pivot_longer(
    cols = c(
      'Energy..Coal...production',
      'Energy..Petroleum...total.petroleum.production',
      'Energy..Natural.gas...production'
    ),
    names_to = "Energy_Source",
    values_to = "Production"
  )

# Filter out the "World" row
long_df <- long_df %>% filter(`Geography..Map.references` != "World")


ggplot(long_df, aes(
  x = `Geography..Map.references`,  # your "continent" column
  y = Production,
  fill = Energy_Source
)) +
  geom_col(position = "dodge") +
  labs(
    x = "Continent",
    y = "Production (BTU)",
    title = "Energy Production by Type and Continent",
    fill = "Energy Source"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


Spider plots
```{r}
# Select relevant columns
energy_df <- wfb_df %>%
  select(
    Continent = `Geography..Map.references`,
    Coal = `Energy..Coal...production`,
    Petroleum = `Energy..Petroleum...total.petroleum.production`,
    Gas = `Energy..Natural.gas...production`
  ) %>%
  group_by(Continent) %>%
  summarise(across(everything(), ~sum(.x, na.rm = TRUE)))

#for each continent
continents <- unique(wfb_df$Geography..Map.references)
continents <- setdiff(continents, "World")

for (continent in continents){
  continent_data <- energy_df %>% filter(Continent == continent) %>% select(-Continent)
  
  # Add max and min rows as required by fmsb
  max_min <- data.frame(
    Coal = c(max(energy_df$Coal), 0),
    Petroleum = c(max(energy_df$Petroleum), 0),
    Gas = c(max(energy_df$Gas), 0)
  )
  
  radar_data <- rbind(max_min, continent_data)
  
  # Plot the radar chart
  radarchart(radar_data, axistype = 1,
             pcol = "blue", pfcol = scales::alpha("blue", 0.3), plwd = 2,
             cglcol = "grey", cglty = 1, axislabcol = "grey", caxislabels = NULL,
             title = paste("Energy Production Profile:", continent))
}

```


```{r}
continent_energy <- wfb_df %>%
  filter(Geography..Map.references != "World") %>%  
  group_by(Geography..Map.references) %>%
  summarise(
    Coal = sum(Energy..Coal...production, na.rm = TRUE),
    Petroleum = sum(Energy..Petroleum...total.petroleum.production, na.rm = TRUE),
    Gas = sum(Energy..Natural.gas...production, na.rm = TRUE)
  )



# Ensure the continent is the row names and transpose the data (this is critical for fmsb)
continent_energy <- as.data.frame(continent_energy)
row.names(continent_energy) <- continent_energy$Geography..Map.references
continent_energy <- continent_energy[, -1]  # Remove the Continent column

normalize <- function(x) (x - min(x)) / (max(x) - min(x))
continent_energy_normalized <- as.data.frame(lapply(continent_energy, normalize))

rownames(continent_energy_normalized) <- rownames(continent_energy)

###individual spider plots by production type
energy_types <- colnames(continent_energy_normalized)
for (type in energy_types) {

  data_vector <- continent_energy_normalized[, type, drop = FALSE]
  
  rownames(data_vector) <- rownames(continent_energy_normalized)

  # Create max-min rows required by fmsb
  radar_data <- rbind(rep(1, nrow(data_vector)), rep(0, nrow(data_vector)), t(data_vector))
  radar_data <- as.data.frame(radar_data)
  # Plot
  radarchart(radar_data,
             axistype = 1,
             title = paste("Normalized", type, "Production by Continent"),
             pcol = "darkgreen", pfcol = scales::alpha("green", 0.3), plwd = 2,
             cglcol = "grey", cglty = 1, axislabcol = "grey")
}

###one plot with each type overlayed
# Transpose the data so that each energy type (Coal, Petroleum, Gas) is a row
continent_energy_normalized <- t(continent_energy_normalized)

# Add required max and min rows for fmsb radar chart (max value and 0)
max_vals <- apply(continent_energy_normalized, 1, max)  # Get max value for each energy type
min_vals <- rep(0, length(max_vals))         # Create a vector of 0s for min values

# Combine max, min, and actual data
radar_data <- rbind(max_vals, min_vals, continent_energy_normalized)

# Convert the matrix to a data frame
radar_data <- as.data.frame(radar_data)

# Plot the radar chart with fmsb
radarchart(radar_data, axistype = 1,
           pcol = c("red", "green", "blue"), 
           pfcol = scales::alpha(c("red", "green", "blue"), 0.3),
           plwd = 2,
           cglcol = "grey", cglty = 1, axislabcol = "grey",
           title = "Energy Production by Continent")

# Add legend
legend("topright", legend = rownames(continent_energy_normalized), 
       col = c("red", "green", "blue"), lty = 1, bty = "n")

```
