---
title: "Question5"
output: html_document
date: "2025-06-18"
---

Can we identify global patterns in agricultural and energy-water usage among countries ? 

Scatter plot: energy imports vs. agricultural yield.

Bar chart: Top 10 countries by energy exports and their food production.


```{r}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(tidyr)
library(ks)
library(rgl)
library(scales)
library(MASS)
library(plotly)
library(fmsb)
library(patchwork)
library(tidyverse)
library(aplpack)
library(leaflet)
library(rnaturalearth)
library(countrycode)
```

```{r}
df <- read.csv("../data/countries.csv")


```


```{r}
bad_values <- c('Akrotiri', 'American Samoa', 'Antarctica', 'Antigua and Barbuda',
                'Ashmore and Cartier Islands', 'Baker Island', 'Bouvet Island',
                'British Indian Ocean Territory', 'British Virgin Islands',
                'Cayman Islands', 'Christmas Island', 'Clipperton Island', 'Cocos (Keeling) Islands',
                'Coral Sea Islands', 'Dhekelia', 'Falkland Islands (Islas Malvinas)', 'Faroe Islands',
                'French Polynesia', 'French Southern and Antarctic Lands', 'Gaza Strip', 'Gibraltar',
                'Guernsey', 'Heard Island and McDonald Islands', 'Howland Island', 'Isle of Man',
                'Jan Mayen', 'Jarvis Island', 'Jersey', 'Johnston Atoll', 'Kingman Reef', 'Macau', 'Midway Islands',
                'Montserrat', 'Navassa Island', 'New Caledonia', 'Niue', 'Norfolk Island', 'Northern Mariana Islands',
                'Palmyra Atoll', 'Paracel Islands', 'Pitcairn Islands', 'Saint Barthelemy', 'Saint Helena, Ascension, and Tristan da Cunha',
                'Saint Martin', 'Saint Pierre and Miquelon', 'Sint Maarten', 'Spratly Islands', 'Svalbard', 'Tokelau',
                'Turks and Caicos Islands', 'Virgin Islands', 'Wake Island', 'Wallis and Futuna', 'World', 'Greenland')

df_clean <- df %>%
  filter(!Country %in% bad_values)

df_clean$Geography..Map.references[df_clean$Geography..Map.references == "Arctic Region"] <- "Europe"
df_clean$Geography..Map.references[df_clean$Geography..Map.references == "AsiaEurope"] <- "Europe"


```


```{r}
# Function to display the columns and nan counts
col_names_and_nas <- function(df) {
  for (i in seq_along(df)) {
    cat(colnames(df)[i], ":", sum(is.na(df[[i]])), "\n")
  }
}

# read data
full_df <- df_clean
```


```{r}
# Function to display the columns and nan counts
col_names_and_nas <- function(df) {
  for (i in seq_along(df)) {
    cat(colnames(df)[i], ":", sum(is.na(df[[i]])), "\n")
  }
}

# Trim data
cols_to_keep <- c("Geography..Map.references","Environment..Total.water.withdrawal...municipal","Environment..Total.water.withdrawal...industrial","Environment..Total.water.withdrawal...agricultural","Environment..Air.pollutants...particulate.matter.emissions","Environment..Air.pollutants...carbon.dioxide.emissions","Environment..Air.pollutants...methane.emissions",                         "Economy..Real.GDP.per.capita", "People.and.Society..Population...total", "People.and.Society..Population.growth.rate", "People.and.Society..Birth.rate", "People.and.Society..Death.rate", "People.and.Society..Maternal.mortality.ratio", "People.and.Society..Infant.mortality.rate...total", "People.and.Society..Life.expectancy.at.birth...total.population" ,"People.and.Society..Physician.density",                   "Energy..Electricity...consumption",
"Energy..Electricity...imports",
"Energy..Coal...production",
"Energy..Coal...consumption",
"Energy..Coal...exports",
"Energy..Coal...imports",
"Energy..Petroleum...refined.petroleum.consumption",
"Energy..Natural.gas...production",
"Energy..Natural.gas...consumption",
"Energy..Petroleum...total.petroleum.production",
"Energy..Petroleum...refined.petroleum.consumption",
"Economy..Average.household.expenditures...on.food",
"Energy..Natural.gas...imports" )

trim_df <- full_df[,cols_to_keep]
trim_df$Country <- full_df[,1]
rownames(trim_df) <- full_df[,1]
trim_df


```


```{r}
library(stringr)

library(dplyr)
library(stringr)

# Geography..Map.references
colnames(trim_df)[colnames(trim_df) == "Geography..Map.references"] <- "Continent"

# Fonction pour extraire, convertir et standardiser les volumes d'eau
convert_water_volume <- function(column) {
  as.numeric(str_extract(column, "[0-9.]+")) -> value
  str_extract(column, "(billion|million|thousand)?\\s*cubic meters") -> unit
  m3 <- case_when(
    str_detect(unit, "billion") ~ value * 1e9,
    str_detect(unit, "million") ~ value * 1e6,
    str_detect(unit, "thousand") ~ value * 1e3,
    TRUE ~ value
  )
  return(m3)
}


# Fonction pour extraire et convertir en metric tons (valeurs numériques)
convert_metric_tons <- function(column) {
  # Remplacer virgules par rien
  clean_col <- str_replace_all(column, ",", "")
  # Extraire la valeur numérique
  value <- as.numeric(str_extract(clean_col, "[0-9.]+"))
  # Extraire le multiplicateur (unit)
  unit <- str_extract(clean_col, "(billion|million|thousand)?\\s*metric tons")
  # Appliquer les conversions en tonnes
  tons <- case_when(
    str_detect(unit, "billion") ~ value * 1e9,
    str_detect(unit, "million") ~ value * 1e6,
    str_detect(unit, "thousand") ~ value * 1e3,
    TRUE ~ value
  )
  return(tons)
}


#Fonction de conversion des barils par jour

convert_bbl_per_day <- function(column) {
  clean_col <- str_replace_all(column, ",", "")
  value <- as.numeric(str_extract(clean_col, "[0-9.]+"))
  unit <- str_extract(clean_col, "(billion|million|thousand)?\\s*bbl/day")

  bbl <- case_when(
    str_detect(unit, "billion") ~ value * 1e9,
    str_detect(unit, "million") ~ value * 1e6,
    str_detect(unit, "thousand") ~ value * 1e3,
    TRUE ~ value
  )
  return(bbl)
}

trim_df <- trim_df %>%
  mutate(
    Water.withdrawal.industrial.m3 = convert_water_volume(`Environment..Total.water.withdrawal...industrial`),
    Water.withdrawal.agricultural.m3 = convert_water_volume(`Environment..Total.water.withdrawal...agricultural`),
    Water.withdrawal.municipal.m3 = convert_water_volume(Environment..Total.water.withdrawal...municipal),
    Natural.gas.imports.m3 = convert_water_volume(Energy..Natural.gas...imports),
    Natural.gas.consumption.m3 = convert_water_volume(Energy..Natural.gas...consumption),
    Natural.gas.production.m3 = convert_water_volume(Energy..Natural.gas...production),
  )

trim_df <- trim_df %>%
  mutate(
    petrol.consumption.bbl = convert_bbl_per_day(Energy..Petroleum...refined.petroleum.consumption),
    petrol.production.bbl = convert_bbl_per_day(Energy..Petroleum...total.petroleum.production)

  )

trim_df <- trim_df %>%
  mutate(
    Coal.consumption.tons = convert_metric_tons(`Energy..Coal...consumption`),
    Coal.exports.tons = convert_metric_tons(`Energy..Coal...exports`),
    Coal.imports.tons = convert_metric_tons(`Energy..Coal...imports`),
    Coal.production.tons = convert_metric_tons(`Energy..Coal...production`)
  )


trim_df <- trim_df[, !(colnames(trim_df) %in% c(
  "Environment..Total.water.withdrawal...industrial",
  "Environment..Total.water.withdrawal...agricultural",
  "Environment..Total.water.withdrawal...municipal",
  "Energy..Natural.gas...imports",
  "Energy..Natural.gas...consumption",
  "Energy..Natural.gas...production",
  "Energy..Coal...consumption",
  "Energy..Coal...exports",
  "Energy..Coal...imports",
  "Energy..Coal...production",
  "Energy..Petroleum...refined.petroleum.consumption",
  "Energy..Petroleum...total.petroleum.production"
  
))]

# Environment: Air pollutants - particulate matter emissions
trim_df$Environment..Air.pollutants...particulate.matter.emissions <- 
  as.numeric(str_extract(trim_df$Environment..Air.pollutants...particulate.matter.emissions, "[0-9.]+"))
colnames(trim_df)[colnames(trim_df) == "Environment..Air.pollutants...particulate.matter.emissions"] <- "particulate_matter_emissions_microgramme_per_cubic_meter"

# Environment: Air pollutants - carbon dioxide emissions
trim_df$Environment..Air.pollutants...carbon.dioxide.emissions <- 
  as.numeric(str_extract(trim_df$Environment..Air.pollutants...carbon.dioxide.emissions, "[0-9.]+"))
colnames(trim_df)[colnames(trim_df) == "Environment..Air.pollutants...carbon.dioxide.emissions"] <- "CO2_emissions_Megatonne"

# Environment: Air pollutants - methane emissions
trim_df$Environment..Air.pollutants...methane.emissions <- 
  as.numeric(str_extract(trim_df$Environment..Air.pollutants...methane.emissions, "[0-9.]+"))
colnames(trim_df)[colnames(trim_df) == "Environment..Air.pollutants...methane.emissions"] <- "methane_emissions_Megatonne"

# Economy..Real.GDP.per.capita
trim_df$Economy..Real.GDP.per.capita <- as.numeric(gsub("[$,]", "", str_extract(trim_df$Economy..Real.GDP.per.capita, "\\$[0-9,]+")))
colnames(trim_df)[colnames(trim_df) == "Economy..Real.GDP.per.capita"] <- "Real.GDP.per.capita.USD"

# People.and.Society..Population.growth.rate
trim_df$People.and.Society..Population.growth.rate <- as.numeric(sub("%", "", str_extract(trim_df$People.and.Society..Population.growth.rate, "[0-9.]+%"))) / 100
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Population.growth.rate"] <- "Population.growth.rate.percent"

# People.and.Society..Birth.rate - births 
trim_df$People.and.Society..Birth.rate <- as.numeric(str_extract(trim_df$People.and.Society..Birth.rate, "^[0-9]+")) / 1000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Birth.rate"] <- "Live.births.per.1000.people"

# People.and.Society..Death.rate - deaths per person 
trim_df$People.and.Society..Death.rate <- as.numeric(str_extract(trim_df$People.and.Society..Death.rate, "^[0-9]+")) / 1000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Death.rate"] <- "Deaths.per.1000.people"

# People.and.Society..Maternal.mortality.ratio - deaths per live birth
trim_df$People.and.Society..Maternal.mortality.ratio <- as.numeric(gsub(",", "", str_extract(trim_df$People.and.Society..Maternal.mortality.ratio, "^[0-9]+"))) / 100000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Maternal.mortality.ratio"] <- "Maternal.deaths.per.100000.live.births"

# People.and.Society..Infant.mortality.rate...total - deaths per live birth
trim_df$People.and.Society..Infant.mortality.rate...total <- as.numeric(str_extract(trim_df$People.and.Society..Infant.mortality.rate...total, "^[0-9.]+")) / 1000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Infant.mortality.rate...total"] <- "Infant.deaths.per.1000.live.births"

# People.and.society..Life.expectancy.at.birth...total.population
trim_df$People.and.Society..Life.expectancy.at.birth...total.population <- as.numeric(str_extract(trim_df$People.and.Society..Life.expectancy.at.birth...total.population, "^[0-9.]+"))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Life.expectancy.at.birth...total.population"] <- "Life.expectancy.at.birth"

# People.and.Society..Population...total
trim_df$People.and.Society..Population...total <- as.numeric(gsub(",", "", str_extract(trim_df$People.and.Society..Population...total, "^[0-9,]+")))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Population...total"] <- "Total.population"

# People.and.Society..Physician.density
trim_df$People.and.Society..Physician.density <- as.numeric(str_extract(trim_df$People.and.Society..Physician.density, "^[0-9.]+")) / 1000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Physician.density"] <- "Physicians.per.1000.people"


# Economy..Average.household.expenditures...on.food
trim_df$Economy..Average.household.expenditures...on.food <- as.numeric(sub("%", "", str_extract(trim_df$Economy..Average.household.expenditures...on.food, "[0-9.]+%"))) / 100
colnames(trim_df)[colnames(trim_df) == "Economy..Average.household.expenditures...on.food"] <- "household.expenditures.food.percent"


trim_df
```



  "Environment..Total.water.withdrawal...industrial",
  "Environment..Total.water.withdrawal...agricultural",
  "Environment..Total.water.withdrawal...municipal",
  "Energy..Natural.gas...imports",
  "Energy..Natural.gas...consumption",
  "Energy..Natural.gas...production",
  "Energy..Coal...consumption",
  "Energy..Coal...exports",
  "Energy..Coal...imports",
  "Energy..Coal...production",
  "Energy..Petroleum...refined.petroleum.consumption",
  "Energy..Petroleum...total.petroleum.production"




```{r}
data1 <- read.csv("../data/FAO_Dataset_19_23.csv", stringsAsFactors = FALSE)
view(data1)
```

```{r}
actual_cols = colnames(data1)
actual_cols
```



```{r}


# 2. Les colonnes que tu veux garder
cols_to_keep <- c(
  "Area",
  "Cereals._primary_Yield_kg.ha",
  "Cereals._primary_Production_t",
  "Eggs_Primary_Yield.Carcass_Weight_g.An",
  "Eggs_Primary_Production_t",
  "Fruit_Primary_Production_t",
  "Fruit_Primary_Yield_kg.ha",
  "Maize_.corn._Yield_kg.ha",
  "Maize_.corn._Production_t",
  "Milk._Total_Production_t",
  "Milk._Total_Milk_Animals_An",
  "Pulses._Total_Production_t",
  "Pulses._Total_Yield_kg.ha",
  "Sheep_and_Goats_Stocks_An",
  "Sheep_and_Goat_Meat_Yield.Carcass_Weight_kg.An",
  "Sugar_Crops_Primary_Yield_kg.ha",
  "Sugar_Crops_Primary_Production_t",
  "Agricultural_land_Share_in_Land_area_.",
  "Agricultural_land_Value_of_agricultural_production_.Int._.._per_Area_USD_PPP.ha",
  "Arable_land_Share_in_Agricultural_land_.",
  "Land_area_equipped_for_irrigation_Share_in_Cropland_.",
  "Permanent_crops_Share_in_Agricultural_land_.",
  "Permanent_meadows_and_pastures_Share_in_Agricultural_land_.",
  "NPK_fertilizers_Export_quantity_t",
  "NPK_fertilizers_Import_quantity_t",
  
  "NPK_fertilizers_Agricultural_Use_t",
  "PK_compounds_Agricultural_Use_t",
  "Urea_Agricultural_Use_t",

  "Herbicides_Agricultural_Use_t",
  "Insecticides_._Biopesticides_Agricultural_Use_t",
  
  "Country_area_Area_1000_ha",                                                                                           "Cropland_Area_1000_ha"
)

# 3. Identifier celles qui manquent
missing_cols <- setdiff(cols_to_keep, actual_cols)

# 4. Afficher les colonnes non trouvées
print("Colonnes non trouvées dans data1 :")
print(missing_cols)

```

```{r}
view(selected_df)
```


```{r}
selected_df <- data1[, cols_to_keep]
```

```{r}
# Noms plus commodes
new_names <- c(
  "Area",
  "cereal_yield_kg_ha",
  "cereal_prod_t",
  "eggs_yield_g_animal",
  "eggs_prod_t",
  "fruit_prod_t",
  "fruit_yield_kg_ha",
  "maize_yield_kg_ha",
  "maize_prod_t",
  "milk_prod_t",
  "milk_animals_count",
  "pulses_prod_t",
  "pulses_yield_kg_ha",
  "sheep_goats_stock",
  "sheep_goat_meat_yield_kg",
  "sugar_crop_yield_kg_ha",
  "sugar_crop_prod_t",
  "agri_land_share_total_land_pct",
  "agri_prod_value_usdppp_per_ha",
  "arable_land_share_agri_pct",
  "irrigation_share_cropland_pct",
  "perm_crops_share_agri_pct",
  "perm_pasture_share_agri_pct",
  "npk_export_t",
  "npk_import_t",
  
  "npk_Agricultural_Use_t",
  "pk_Agricultural_Use_t",
  "Urea_Agricultural_Use_t",
  

  
  "Herbicides_Agricultural_Use_t",
  "Insecticides_Agricultural_Use_t",
  "Country_area1000_ha",                                                                                                "Cropland_Area_1000_ha"
)

# Renommer
colnames(selected_df) <- new_names

data1 = selected_df
data1$Area[222] <- "United States"
data1$Area[216] <- "Turkey (Turkiye)"

data1$Area[220] <- "United Kingdom"

```

```{r}
trim_df$ID_Npm_Country <- rownames(trim_df)

```


```{r}
merged_df <- merge(trim_df, data1, 
                   by.x = "ID_Npm_Country",   # colonne dans trim_df
                   by.y = "Area",     # colonne correspondante dans data1
                   all.x = TRUE)              # merge à gauche (keep all trim_df rows)
merged_df <- merged_df %>%
  mutate(npk_agricultural_use_ha = npk_Agricultural_Use_t / (Cropland_Area_1000_ha * 1000))
merged_df
```










# overview



```{r}
library(ggplot2)


ggplot(merged_df, aes(x = petrol.consumption.bbl, y = cereal_yield_kg_ha)) +
  geom_point(aes(color = Continent), size = 3, alpha = 0.7) +
  scale_x_log10() +  # better spread
  labs(
    title = "petrol consumption vs Cereal Yield",
    x = "petrol consumption bbl (log scale)",
    y = "Cereal Yield (kg/ha)"
  ) +
  theme_minimal()
```
# PCA 

```{r}
# Chargement des bibliothèques
library(tidyverse)
library(FactoMineR)
library(factoextra)
library(janitor)

# Variables retenues
Vars_Simple <- c(
  "cereal_yield_kg_ha",  
  "npk_import_t",
  "npk_export_t",
  "Natural.gas.imports.m3", 
  "Coal.imports.tons", 

  "Water.withdrawal.agricultural.m3",
  "methane_emissions_Megatonne"
)

  
vars_pca <- c(
  "cereal_yield_kg_ha", 
  "fruit_yield_kg_ha", 
  "maize_yield_kg_ha",
  "pulses_yield_kg_ha",
  
  #"sugar_crop_yield_kg_ha", 
  #"eggs_yield_g_animal",
  #"sheep_goat_meat_yield_kg",

  #"npk_import_t", 
  #"npk_export_t",
  

  "Coal.consumption.tons",
  "petrol.consumption.bbl",
  "Natural.gas.consumption.m3",

  
  
  "npk_Agricultural_Use_t",
  "npk_agricultural_use_ha",
  #"pk_Agricultural_Use_t",
  #"Urea_Agricultural_Use_t",
  #"Herbicides_Agricultural_Use_t",
  #"Insecticides_Agricultural_Use_t",

  "Water.withdrawal.agricultural.m3",

  #"agri_land_share_total_land_pct", 
  #"arable_land_share_agri_pct",
  #"perm_crops_share_agri_pct", 
  #"perm_pasture_share_agri_pct",
  #"irrigation_share_cropland_pct"

  #"agri_prod_value_usdppp_per_ha"
  "Country_area1000_ha",                                                                                                "Cropland_Area_1000_ha",
 
 # Production 
 
  "cereal_prod_t",
  "eggs_prod_t",
  "fruit_prod_t",
  "maize_prod_t",
  "milk_prod_t",
  "pulses_prod_t",
  "sugar_crop_prod_t"
)

var_categories <- c(

  # Yield (Production efficiency)
  "cereal_yield_kg_ha" = "Yield",
  "fruit_yield_kg_ha" = "Yield",
  "maize_yield_kg_ha" = "Yield",
  "pulses_yield_kg_ha" = "Yield",
  "sugar_crop_yield_kg_ha" = "Yield",
  "eggs_yield_g_animal" = "Yield",
  "sheep_goat_meat_yield_kg" = "Yield",

  # Inputs: Fertilizer / Energy

  
  "Coal.consumption.tons"= "Energy",
  "petrol.consumption.bbl"= "Energy",
  "Natural.gas.consumption.m3"= "Energy",
  "npk_import_t" = "Fertilizer",
  "npk_export_t" = "Fertilizer",
  "npk_agricultural_use_ha"= "Fertilizer",
  "npk_Agricultural_Use_t"= "Fertilizer",
  "pk_Agricultural_Use_t"= "Fertilizer",
  "Urea_Agricultural_Use_t"= "Fertilizer",
  "Herbicides_Agricultural_Use_t"= "Fertilizer",
  "Insecticides_Agricultural_Use_t"= "Fertilizer",

  # Water
  "Water.withdrawal.agricultural.m3" = "Water",

  # Agricultural structure (% land use, irrigation)
  "agri_land_share_total_land_pct" = "Agri_%_Structure",
  "arable_land_share_agri_pct" = "Agri_%_Structure",
  "perm_crops_share_agri_pct" = "Agri_%_Structure",
  "perm_pasture_share_agri_pct" = "Agri_%_Structure",
  "irrigation_share_cropland_pct" = "Agri_%_Structure",

  # Value
  "agri_prod_value_usdppp_per_ha" = "Agri_Value",
  "Country_area1000_ha" = "Agri_Value",                                                                                 "Cropland_Area_1000_ha" = "Agri_Value",
 
 # Production 
  "cereal_prod_t"="Production",
  "eggs_prod_t"="Production",
  "fruit_prod_t"="Production",
  "maize_prod_t"="Production",
  "milk_prod_t"="Production",
  "pulses_prod_t"="Production",
  "sugar_crop_prod_t"="Production"
)



```

```{r}
# 1. Identifier les pays avec ≥ 1% de la production totale
countries_above_threshold <- merged_df %>%
  select(ID_Npm_Country, cereal_prod_t) %>%
  filter(!is.na(cereal_prod_t)) %>%
  group_by(ID_Npm_Country) %>%
  summarise(total_prod = sum(cereal_prod_t, na.rm = TRUE)) %>%
  mutate(perc = 100 * total_prod / sum(total_prod)) %>%
  filter(perc >= 1) %>%
  pull(ID_Npm_Country)

# 2. Construire pca_data UNIQUEMENT avec ces pays
filter_data <- merged_df %>%
  filter(ID_Npm_Country %in% countries_above_threshold)


pca_data <- merged_df %>%
  select(all_of(vars_pca), Continent, ID_Npm_Country) %>%
  drop_na()  # tu peux aussi faire imput si trop restrictif

pca_data <- merged_df %>%
  select(all_of(vars_pca), Continent, ID_Npm_Country) %>%
  filter(rowMeans(is.na(.)) <= 0.2)



library(missMDA)
X_incomplete <- scale(pca_data %>% select(all_of(vars_pca)))


# Imputation
imputed <- imputePCA(X_incomplete, ncp = 2)  # ou plus de dimensions
X <- imputed$completeObs


# 3. Supprimer les colonnes constantes et standardiser
# Standardiser les données (important pour la PCA)


```

```{r}
pca_data
```







```{r}
library(dplyr)
library(ggplot2)
library(scales)

# 1. Ajouter le rendement (par moyenne si plusieurs lignes par pays)
yield_df <- merged_df %>%
  select(ID_Npm_Country, cereal_yield_kg_ha) %>%
  group_by(ID_Npm_Country) %>%
  summarise(mean_yield = mean(cereal_yield_kg_ha, na.rm = TRUE))

# 2. Rejoindre avec df_plot
df_plot <- df_plot %>%
  left_join(yield_df, by = "ID_Npm_Country")

# 3. Barplot avec couleur = rendement
ggplot(df_plot, aes(x = reorder(ID_Npm_Country, -perc), y = perc, fill = mean_yield)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Cereal production per country",
    x = "Country",
    y = "Production (%)",
    fill = "Yield\nmean (kg/ha)"
  ) +
  scale_y_continuous(labels = percent_format(scale = 1)) +
  scale_fill_viridis_c(option = "D", direction = -1) +  # ou scale_fill_gradient()
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

```




```{r}
res_pca <- prcomp(X, center = TRUE, scale. = TRUE)

```


```{r}
fviz_eig(res_pca, addlabels = TRUE, ylim = c(0, 50)) +
  ggtitle("Scree Plot: Variance Explained by Principal Components") +
  theme_minimal()
```


```{r}
fviz_pca_var(res_pca,
             col.var = var_categories[colnames(X)],
             #col.var = "contrib",
             palette = "Set2",
             repel = TRUE) +
  theme_minimal()
```


```{r}
fviz_contrib(res_pca, choice = "var", axes = 1, top = 15)
fviz_contrib(res_pca, choice = "var", axes = 2, top = 15)
```


```{r}
fviz_pca_var(res_pca,
             col.var = "contrib",          # couleur par contribution
             gradient.cols = c("#00AFBB", "#E7B800", "#FC4E07"),
             repel = TRUE) +
  ggtitle("Correlation Circle (Variables)") +
  theme_minimal()
```

```{r}
fviz_pca_ind(res_pca,
             geom = "point",
             col.ind = pca_data$Continent,
             palette = "jco",
             addEllipses = TRUE,
             legend.title = "Continent",
             repel = TRUE) +
  ggtitle("Individuals Factor Map (colored by Continent)") +
  theme_minimal()
```
```{r}
library(ggrepel)

library(ggplot2)
library(ggrepel)

# Extraire les scores PCA
ind_coords <- as.data.frame(res_pca$x[, 1:2])
ind_coords$Country <- pca_data$ID_Npm_Country

# Plot
ggplot(ind_coords, aes(x = PC1, y = PC2, label = Country)) +
  geom_point(color = "steelblue") +
  geom_text_repel(size = 3, max.overlaps = 100) +
  theme_minimal() +
  ggtitle("Individuals Factor Map (labels instead of legend)")

```

```{r}
fviz_pca_biplot(res_pca, 
                label = "var", 
                habillage = pca_data$Continent, 
                palette = "jco",
                addEllipses = TRUE, 
                repel = TRUE, 
                col.var = "steelblue")
```

```{r}
library(corrplot)
corr_data <- merged_df %>% select(all_of(vars_pca)) %>% cor(use = "complete.obs")
corrplot(corr_data, method = "color", type = "upper", tl.col = "black", addCoef.col = "black")

```


```{r}
library(randomForest)
rf <- randomForest(cereal_yield_kg_ha ~ ., data = X, importance = TRUE)
varImpPlot(rf)
```

