---
title: "Question5"
output:
  pdf_document: default
  html_document: default
date: "2025-06-18"
---

# Can we identify global patterns in agricultural performance and energy-water usage among countries ? 




```{r,include=FALSE}
library(dplyr)
library(ggplot2)
library(ggrepel)
library(tidyr)
library(ks)
library(rgl)
library(scales)
library(MASS)
library(plotly)
library(fmsb)
library(patchwork)
library(tidyverse)
library(aplpack)
library(leaflet)
library(rnaturalearth)
library(countrycode)
library(GGally)
library(FactoMineR)
library(factoextra)
library(janitor)



df <- read.csv("../data/countries.csv")


bad_values <- c('Akrotiri', 'American Samoa', 'Antarctica', 'Antigua and Barbuda',
                'Ashmore and Cartier Islands', 'Baker Island', 'Bouvet Island',
                'British Indian Ocean Territory', 'British Virgin Islands',
                'Cayman Islands', 'Christmas Island', 'Clipperton Island', 'Cocos (Keeling) Islands',
                'Coral Sea Islands', 'Dhekelia', 'Falkland Islands (Islas Malvinas)', 'Faroe Islands',
                'French Polynesia', 'French Southern and Antarctic Lands', 'Gaza Strip', 'Gibraltar',
                'Guernsey', 'Heard Island and McDonald Islands', 'Howland Island', 'Isle of Man',
                'Jan Mayen', 'Jarvis Island', 'Jersey', 'Johnston Atoll', 'Kingman Reef', 'Macau', 'Midway Islands',
                'Montserrat', 'Navassa Island', 'New Caledonia', 'Niue', 'Norfolk Island', 'Northern Mariana Islands',
                'Palmyra Atoll', 'Paracel Islands', 'Pitcairn Islands', 'Saint Barthelemy', 'Saint Helena, Ascension, and Tristan da Cunha',
                'Saint Martin', 'Saint Pierre and Miquelon', 'Sint Maarten', 'Spratly Islands', 'Svalbard', 'Tokelau',
                'Turks and Caicos Islands', 'Virgin Islands', 'Wake Island', 'Wallis and Futuna', 'World', 'Greenland')

df_clean <- df %>%
  filter(!Country %in% bad_values)

df_clean$Geography..Map.references[df_clean$Geography..Map.references == "Arctic Region"] <- "Europe"
df_clean$Geography..Map.references[df_clean$Geography..Map.references == "AsiaEurope"] <- "Europe"



# Function to display the columns and nan counts
col_names_and_nas <- function(df) {
  for (i in seq_along(df)) {
    cat(colnames(df)[i], ":", sum(is.na(df[[i]])), "\n")
  }
}

# read data
full_df <- df_clean

# Function to display the columns and nan counts
col_names_and_nas <- function(df) {
  for (i in seq_along(df)) {
    cat(colnames(df)[i], ":", sum(is.na(df[[i]])), "\n")
  }
}

# Trim data
cols_to_keep <- c("Geography..Map.references","Environment..Total.water.withdrawal...municipal","Environment..Total.water.withdrawal...industrial","Environment..Total.water.withdrawal...agricultural","Environment..Air.pollutants...particulate.matter.emissions","Environment..Air.pollutants...carbon.dioxide.emissions","Environment..Air.pollutants...methane.emissions",                         "Economy..Real.GDP.per.capita", "People.and.Society..Population...total", "People.and.Society..Population.growth.rate", "People.and.Society..Birth.rate", "People.and.Society..Death.rate", "People.and.Society..Maternal.mortality.ratio", "People.and.Society..Infant.mortality.rate...total", "People.and.Society..Life.expectancy.at.birth...total.population" ,"People.and.Society..Physician.density",                   "Energy..Electricity...consumption",
"Energy..Electricity...imports",
"Energy..Coal...production",
"Energy..Coal...consumption",
"Energy..Coal...exports",
"Energy..Coal...imports",
"Energy..Petroleum...refined.petroleum.consumption",
"Energy..Natural.gas...production",
"Energy..Natural.gas...consumption",
"Energy..Petroleum...total.petroleum.production",
"Energy..Petroleum...refined.petroleum.consumption",
"Economy..Average.household.expenditures...on.food",
"Energy..Natural.gas...imports" )

trim_df <- full_df[,cols_to_keep]
trim_df$Country <- full_df[,1]
rownames(trim_df) <- full_df[,1]



library(dplyr)
library(stringr)

# Geography..Map.references
colnames(trim_df)[colnames(trim_df) == "Geography..Map.references"] <- "Continent"

# Fonction pour extraire, convertir et standardiser les volumes d'eau
convert_water_volume <- function(column) {
  as.numeric(str_extract(column, "[0-9.]+")) -> value
  str_extract(column, "(billion|million|thousand)?\\s*cubic meters") -> unit
  m3 <- case_when(
    str_detect(unit, "billion") ~ value * 1e9,
    str_detect(unit, "million") ~ value * 1e6,
    str_detect(unit, "thousand") ~ value * 1e3,
    TRUE ~ value
  )
  return(m3)
}


# Fonction pour extraire et convertir en metric tons (valeurs numériques)
convert_metric_tons <- function(column) {
  # Remplacer virgules par rien
  clean_col <- str_replace_all(column, ",", "")
  # Extraire la valeur numérique
  value <- as.numeric(str_extract(clean_col, "[0-9.]+"))
  # Extraire le multiplicateur (unit)
  unit <- str_extract(clean_col, "(billion|million|thousand)?\\s*metric tons")
  # Appliquer les conversions en tonnes
  tons <- case_when(
    str_detect(unit, "billion") ~ value * 1e9,
    str_detect(unit, "million") ~ value * 1e6,
    str_detect(unit, "thousand") ~ value * 1e3,
    TRUE ~ value
  )
  return(tons)
}


#Fonction de conversion des barils par jour

convert_bbl_per_day <- function(column) {
  clean_col <- str_replace_all(column, ",", "")
  value <- as.numeric(str_extract(clean_col, "[0-9.]+"))
  unit <- str_extract(clean_col, "(billion|million|thousand)?\\s*bbl/day")

  bbl <- case_when(
    str_detect(unit, "billion") ~ value * 1e9,
    str_detect(unit, "million") ~ value * 1e6,
    str_detect(unit, "thousand") ~ value * 1e3,
    TRUE ~ value
  )
  return(bbl)
}

trim_df <- trim_df %>%
  mutate(
    Water.withdrawal.industrial.m3 = convert_water_volume(`Environment..Total.water.withdrawal...industrial`),
    Water.withdrawal.agricultural.m3 = convert_water_volume(`Environment..Total.water.withdrawal...agricultural`),
    Water.withdrawal.municipal.m3 = convert_water_volume(Environment..Total.water.withdrawal...municipal),
    Natural.gas.imports.m3 = convert_water_volume(Energy..Natural.gas...imports),
    Natural.gas.consumption.m3 = convert_water_volume(Energy..Natural.gas...consumption),
    Natural.gas.production.m3 = convert_water_volume(Energy..Natural.gas...production),
  )

trim_df <- trim_df %>%
  mutate(
    petrol.consumption.bbl = convert_bbl_per_day(Energy..Petroleum...refined.petroleum.consumption),
    petrol.production.bbl = convert_bbl_per_day(Energy..Petroleum...total.petroleum.production)

  )

trim_df <- trim_df %>%
  mutate(
    Coal.consumption.tons = convert_metric_tons(`Energy..Coal...consumption`),
    Coal.exports.tons = convert_metric_tons(`Energy..Coal...exports`),
    Coal.imports.tons = convert_metric_tons(`Energy..Coal...imports`),
    Coal.production.tons = convert_metric_tons(`Energy..Coal...production`)
  )


trim_df <- trim_df[, !(colnames(trim_df) %in% c(
  "Environment..Total.water.withdrawal...industrial",
  "Environment..Total.water.withdrawal...agricultural",
  "Environment..Total.water.withdrawal...municipal",
  "Energy..Natural.gas...imports",
  "Energy..Natural.gas...consumption",
  "Energy..Natural.gas...production",
  "Energy..Coal...consumption",
  "Energy..Coal...exports",
  "Energy..Coal...imports",
  "Energy..Coal...production",
  "Energy..Petroleum...refined.petroleum.consumption",
  "Energy..Petroleum...total.petroleum.production"
  
))]

# Environment: Air pollutants - particulate matter emissions
trim_df$Environment..Air.pollutants...particulate.matter.emissions <- 
  as.numeric(str_extract(trim_df$Environment..Air.pollutants...particulate.matter.emissions, "[0-9.]+"))
colnames(trim_df)[colnames(trim_df) == "Environment..Air.pollutants...particulate.matter.emissions"] <- "particulate_matter_emissions_microgramme_per_cubic_meter"

# Environment: Air pollutants - carbon dioxide emissions
trim_df$Environment..Air.pollutants...carbon.dioxide.emissions <- 
  as.numeric(str_extract(trim_df$Environment..Air.pollutants...carbon.dioxide.emissions, "[0-9.]+"))
colnames(trim_df)[colnames(trim_df) == "Environment..Air.pollutants...carbon.dioxide.emissions"] <- "CO2_emissions_Megatonne"

# Environment: Air pollutants - methane emissions
trim_df$Environment..Air.pollutants...methane.emissions <- 
  as.numeric(str_extract(trim_df$Environment..Air.pollutants...methane.emissions, "[0-9.]+"))
colnames(trim_df)[colnames(trim_df) == "Environment..Air.pollutants...methane.emissions"] <- "methane_emissions_Megatonne"

# Economy..Real.GDP.per.capita
trim_df$Economy..Real.GDP.per.capita <- as.numeric(gsub("[$,]", "", str_extract(trim_df$Economy..Real.GDP.per.capita, "\\$[0-9,]+")))
colnames(trim_df)[colnames(trim_df) == "Economy..Real.GDP.per.capita"] <- "Real.GDP.per.capita.USD"

# People.and.Society..Population.growth.rate
trim_df$People.and.Society..Population.growth.rate <- as.numeric(sub("%", "", str_extract(trim_df$People.and.Society..Population.growth.rate, "[0-9.]+%"))) / 100
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Population.growth.rate"] <- "Population.growth.rate.percent"

# People.and.Society..Birth.rate - births 
trim_df$People.and.Society..Birth.rate <- as.numeric(str_extract(trim_df$People.and.Society..Birth.rate, "^[0-9]+")) / 1000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Birth.rate"] <- "Live.births.per.1000.people"

# People.and.Society..Death.rate - deaths per person 
trim_df$People.and.Society..Death.rate <- as.numeric(str_extract(trim_df$People.and.Society..Death.rate, "^[0-9]+")) / 1000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Death.rate"] <- "Deaths.per.1000.people"

# People.and.Society..Maternal.mortality.ratio - deaths per live birth
trim_df$People.and.Society..Maternal.mortality.ratio <- as.numeric(gsub(",", "", str_extract(trim_df$People.and.Society..Maternal.mortality.ratio, "^[0-9]+"))) / 100000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Maternal.mortality.ratio"] <- "Maternal.deaths.per.100000.live.births"

# People.and.Society..Infant.mortality.rate...total - deaths per live birth
trim_df$People.and.Society..Infant.mortality.rate...total <- as.numeric(str_extract(trim_df$People.and.Society..Infant.mortality.rate...total, "^[0-9.]+")) / 1000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Infant.mortality.rate...total"] <- "Infant.deaths.per.1000.live.births"

# People.and.society..Life.expectancy.at.birth...total.population
trim_df$People.and.Society..Life.expectancy.at.birth...total.population <- as.numeric(str_extract(trim_df$People.and.Society..Life.expectancy.at.birth...total.population, "^[0-9.]+"))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Life.expectancy.at.birth...total.population"] <- "Life.expectancy.at.birth"

# People.and.Society..Population...total
trim_df$People.and.Society..Population...total <- as.numeric(gsub(",", "", str_extract(trim_df$People.and.Society..Population...total, "^[0-9,]+")))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Population...total"] <- "Total.population"

# People.and.Society..Physician.density
trim_df$People.and.Society..Physician.density <- as.numeric(str_extract(trim_df$People.and.Society..Physician.density, "^[0-9.]+")) / 1000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Physician.density"] <- "Physicians.per.1000.people"


# Economy..Average.household.expenditures...on.food
trim_df$Economy..Average.household.expenditures...on.food <- as.numeric(sub("%", "", str_extract(trim_df$Economy..Average.household.expenditures...on.food, "[0-9.]+%"))) / 100
colnames(trim_df)[colnames(trim_df) == "Economy..Average.household.expenditures...on.food"] <- "household.expenditures.food.percent"


data1 <- read.csv("../data/FAO_Dataset_19_23.csv", stringsAsFactors = FALSE)


# 2. Les colonnes que tu veux garder
cols_to_keep <- c(
  "Area",
  "Cereals._primary_Yield_kg.ha",
  "Cereals._primary_Production_t",
  "Eggs_Primary_Yield.Carcass_Weight_g.An",
  "Eggs_Primary_Production_t",
  "Fruit_Primary_Production_t",
  "Fruit_Primary_Yield_kg.ha",
  "Maize_.corn._Yield_kg.ha",
  "Maize_.corn._Production_t",
  "Milk._Total_Production_t",
  "Milk._Total_Milk_Animals_An",
  "Pulses._Total_Production_t",
  "Pulses._Total_Yield_kg.ha",
  "Sheep_and_Goats_Stocks_An",
  "Sheep_and_Goat_Meat_Yield.Carcass_Weight_kg.An",
  "Sugar_Crops_Primary_Yield_kg.ha",
  "Sugar_Crops_Primary_Production_t",
  "Agricultural_land_Share_in_Land_area_.",
  "Agricultural_land_Value_of_agricultural_production_.Int._.._per_Area_USD_PPP.ha",
  "Arable_land_Share_in_Agricultural_land_.",
  "Land_area_equipped_for_irrigation_Share_in_Cropland_.",
  "Permanent_crops_Share_in_Agricultural_land_.",
  "Permanent_meadows_and_pastures_Share_in_Agricultural_land_.",
  "NPK_fertilizers_Export_quantity_t",
  "NPK_fertilizers_Import_quantity_t",
  
  "NPK_fertilizers_Agricultural_Use_t",
  "PK_compounds_Agricultural_Use_t",
  "Urea_Agricultural_Use_t",

  "Herbicides_Agricultural_Use_t",
  "Insecticides_._Biopesticides_Agricultural_Use_t",
  
  "Country_area_Area_1000_ha",                                                                                           "Cropland_Area_1000_ha"
)



selected_df <- data1[, cols_to_keep]

# Noms plus commodes
new_names <- c(
  "Area",
  "cereal_yield_kg_ha",
  "cereal_prod_t",
  "eggs_yield_g_animal",
  "eggs_prod_t",
  "fruit_prod_t",
  "fruit_yield_kg_ha",
  "maize_yield_kg_ha",
  "maize_prod_t",
  "milk_prod_t",
  "milk_animals_count",
  "pulses_prod_t",
  "pulses_yield_kg_ha",
  "sheep_goats_stock",
  "sheep_goat_meat_yield_kg",
  "sugar_crop_yield_kg_ha",
  "sugar_crop_prod_t",
  "agri_land_share_total_land_pct",
  "agri_prod_value_usdppp_per_ha",
  "arable_land_share_agri_pct",
  "irrigation_share_cropland_pct",
  "perm_crops_share_agri_pct",
  "perm_pasture_share_agri_pct",
  "npk_export_t",
  "npk_import_t",
  
  "npk_Agricultural_Use_t",
  "pk_Agricultural_Use_t",
  "Urea_Agricultural_Use_t",
  

  
  "Herbicides_Agricultural_Use_t",
  "Insecticides_Agricultural_Use_t",
  "Country_area1000_ha",                                                                                                "Cropland_Area_1000_ha"
)

# Renommer
colnames(selected_df) <- new_names

data1 = selected_df
data1$Area[222] <- "United States"
data1$Area[216] <- "Turkey (Turkiye)"
data1$Area[174] <-  "Russia" 

data1$Area[220] <- "United Kingdom"

trim_df$ID_Npm_Country <- rownames(trim_df)

merged_df <- merge(trim_df, data1, 
                   by.x = "ID_Npm_Country",   # colonne dans trim_df
                   by.y = "Area",     # colonne correspondante dans data1
                   all.x = TRUE)              # merge à gauche (keep all trim_df rows)
merged_df <- merged_df %>%
  mutate(npk_per_ha = npk_Agricultural_Use_t / (Cropland_Area_1000_ha * 1000))

merged_df <- merged_df %>%
  mutate(Herbicides_per_HA = Herbicides_Agricultural_Use_t / (Cropland_Area_1000_ha * 1000))

merged_df <- merged_df %>%
  mutate(Insecticides_per_HA = Insecticides_Agricultural_Use_t / (Cropland_Area_1000_ha * 1000))

merged_df <- merged_df %>%
  mutate(Coal.consumption.per.HA = Coal.consumption.tons / (Cropland_Area_1000_ha * 1000))

merged_df <- merged_df %>%
  mutate(petrol.consumption.per.HA = petrol.consumption.bbl / (Cropland_Area_1000_ha * 1000))

merged_df <- merged_df %>%
  mutate(Natural.gas.per.HA = Natural.gas.consumption.m3 / (Cropland_Area_1000_ha * 1000))

merged_df <- merged_df %>%
  mutate(agricultural.Water.withdrawal.per.HA = Water.withdrawal.agricultural.m3 / (Cropland_Area_1000_ha * 1000))

remove_outliers_iqr <- function(df, cols) {
  for (col in cols) {
    Q1 <- quantile(df[[col]], 0.25, na.rm = TRUE)
    Q3 <- quantile(df[[col]], 0.75, na.rm = TRUE)
    IQR_val <- Q3 - Q1
    lower_bound <- Q1 - 1.5 * IQR_val
    upper_bound <- Q3 + 1.5 * IQR_val
    
    df <- df %>% filter(.data[[col]] >= lower_bound & .data[[col]] <= upper_bound)
  }
  return(df)
}

# Liste des colonnes pour lesquelles on veut retirer les outliers
vars_to_clean <- c(
"cereal_yield_kg_ha"
)

# Appliquer le filtre
merged_df <- remove_outliers_iqr(merged_df, vars_to_clean)



merged_df$Continent[51] <-  "Europe" 


```








# overview



```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(ggplot2)


# Graphique
ggplot(merged_df, aes(x = petrol.consumption.per.HA, y = cereal_yield_kg_ha)) +
  geom_point(aes(color = Continent, size = merged_df$cereal_prod_t), alpha = 0.7) +
  scale_x_log10() +
  scale_size_continuous(name = "Total Cereal Production (T)", range = c(1, 10)) +
  labs(
    title = "Petrol Consumption vs Cereal Yield",
    x = "Petrol Consumption per ha (log scale)",
    y = "Cereal Yield (kg/ha)"
  ) +
  theme_minimal()
```
This scatter plot suggests a potential positive correlation between petrol consumption per hectare and cereal yield. Additionally, countries with higher total cereal production (indicated by the size of the points) also tend to exhibit higher petrol usage and yield levels.

A plausible hypothesis emerging from this visualization is that countries with higher energy inputs may achieve more productive agricultural systems. 



```{r, echo= FALSE, warning=FALSE, message=FALSE}
library(ggplot2)
library(viridis)
library(hrbrthemes)  # pour theme_ipsum
library(corrplot)
library( missMDA)
library(ggplot2)
library(viridis)


ggplot(merged_df, aes(x = cereal_yield_kg_ha, fill = Continent)) +
  geom_density(alpha = 0.7, color = NA) +  # Pas de contour = plus clair
  scale_fill_viridis(discrete = TRUE, option = "D") +
  facet_wrap(~ Continent, scales = "free_y") +  # Un graphe par continent
  theme_minimal(base_size = 14) +  # Simplicité + lisibilité
  theme(
    legend.position = "none",  # Pas besoin de légende
    strip.text = element_text(size = 12, face = "bold")
  ) +
  xlab("Cereal Yield (kg/ha)") +
  ylab("Density")

```

```{r, include=FALSE}

vars_corr <- c(


  
  "Coal.consumption.per.HA",
  "petrol.consumption.per.HA",
  "Natural.gas.per.HA",
  "agricultural.Water.withdrawal.per.HA",
  

  
  #"Insecticides_per_HA",
  "npk_per_ha",
  #"Herbicides_per_HA",
  "agri_prod_value_usdppp_per_ha",

  
  "Coal.consumption.tons",
  "petrol.consumption.bbl",
  "Natural.gas.consumption.m3",
  "Water.withdrawal.agricultural.m3",
  
  #"npk_Agricultural_Use_t",
  #"pk_Agricultural_Use_t",
  #"Urea_Agricultural_Use_t",
  #"Herbicides_Agricultural_Use_t",
  #"Insecticides_Agricultural_Use_t",
  
  #"agri_land_share_total_land_pct", 
  #"arable_land_share_agri_pct",
  #"perm_crops_share_agri_pct", 
  #"perm_pasture_share_agri_pct",
  #"irrigation_share_cropland_pct"


  #"Country_area1000_ha",                                                                                               # "Cropland_Area_1000_ha"
 
 # Production 
 
  "cereal_prod_t",
  #"eggs_prod_t",
  #"fruit_prod_t",
  "maize_prod_t",
  "milk_prod_t",
  "pulses_prod_t"
  #"sugar_crop_prod_t"
)
corr_data <- merged_df %>% select(all_of(vars_corr))
# Imputation
imputed <- imputePCA(corr_data, ncp = 2)
corr_data <- imputed$completeObs

corr_data <- as.data.frame(corr_data)
corr_data$Continent <- merged_df$Continent
corr_data$Country <- merged_df$ID_Npm_Country

#corrplot(corr_data, method = "color", type = "upper", tl.col = "black", addCoef.col = "black")
```




# Principal Component Analysis (PCA)
In this analysis, we performed a Principal Component Analysis (PCA) on a set of standardized (per hectare of arable land) agricultural, environmental, and input-related variables across countries. T
he variables include several yield measures (e.g., cereal, maize, fruit, pulses, sugar crops), energy consumption per hectare (coal, petrol, natural gas), agricultural water withdrawal per hectare, fertilizer input (NPK, herbicides, insecticides), agricultural production value per hectare, and the share of irrigated cropland.

These variables were grouped into thematic categories—Yield, Energy, Fertilizer, Water, Agricultural Structure, Value, and Production—to better understand the multidimensional structure of agricultural efficiency and resource use across countries.

```{r,include=FALSE}
# Chargement des bibliothèques



  
vars_pca_per_HA <- c(
  "cereal_yield_kg_ha", 
  "fruit_yield_kg_ha", 
  "maize_yield_kg_ha",
  "pulses_yield_kg_ha",
  "sugar_crop_yield_kg_ha", 
  #"eggs_yield_g_animal",
  #"sheep_goat_meat_yield_kg",

  
  "Coal.consumption.per.HA",
  "petrol.consumption.per.HA",
  "Natural.gas.per.HA",
  "agricultural.Water.withdrawal.per.HA",
  

  
  "Insecticides_per_HA",
  "npk_per_ha",
  "Herbicides_per_HA",
  "agri_prod_value_usdppp_per_ha",


  "irrigation_share_cropland_pct"
)

vars_pca_abso <- c(

  "agri_prod_value_usdppp_per_ha",

  
  "Coal.consumption.tons",
  "petrol.consumption.bbl",
  "Natural.gas.consumption.m3",
  "Water.withdrawal.agricultural.m3",
  
  "npk_Agricultural_Use_t",
  "pk_Agricultural_Use_t",
  "Urea_Agricultural_Use_t",
  "Herbicides_Agricultural_Use_t",
  "Insecticides_Agricultural_Use_t",
  

  "Country_area1000_ha",                                                                                               #             
  "Cropland_Area_1000_ha",
 
 # Production 
 
  "cereal_prod_t",
  "eggs_prod_t",
  "fruit_prod_t",
  "maize_prod_t",
  "milk_prod_t",
  "pulses_prod_t",
  "sugar_crop_prod_t"
)

var_categories <- c(

  # Yield (Production efficiency)
  "cereal_yield_kg_ha" = "Yield",
  "fruit_yield_kg_ha" = "Yield",
  "maize_yield_kg_ha" = "Yield",
  "pulses_yield_kg_ha" = "Yield",
  "sugar_crop_yield_kg_ha" = "Yield",
  #"eggs_yield_g_animal" = "Yield",
  #"sheep_goat_meat_yield_kg" = "Yield",

  # Inputs: Fertilizer / Energy

  
  "Coal.consumption.tons"= "Energy",
  "petrol.consumption.bbl"= "Energy",
  "Natural.gas.consumption.m3"= "Energy",
    
  "Coal.consumption.per.HA"= "Energy",
  "petrol.consumption.per.HA"= "Energy",
  "Natural.gas.per.HA"= "Energy",
  "agricultural.Water.withdrawal.per.HA"= "Energy",
  
  
  
  "Insecticides_per_HA" = "Fertilizer",
  
  "Herbicides_per_HA" = "Fertilizer",
  "npk_per_ha"= "Fertilizer",
  "npk_Agricultural_Use_t"= "Fertilizer",
  "pk_Agricultural_Use_t"= "Fertilizer",
  "Urea_Agricultural_Use_t"= "Fertilizer",
  "Herbicides_Agricultural_Use_t"= "Fertilizer",
  "Insecticides_Agricultural_Use_t"= "Fertilizer",

  # Water
  "Water.withdrawal.agricultural.m3" = "Water",

  # Agricultural structure (% land use, irrigation)
  "agri_land_share_total_land_pct" = "Agri_%_Structure",
  "arable_land_share_agri_pct" = "Agri_%_Structure",
  "perm_crops_share_agri_pct" = "Agri_%_Structure",
  "perm_pasture_share_agri_pct" = "Agri_%_Structure",
  "irrigation_share_cropland_pct" = "Agri_%_Structure",

  # Value
  "agri_prod_value_usdppp_per_ha" = "Agri_Value",
  "Country_area1000_ha" = "Agri_Value",                                                                                 "Cropland_Area_1000_ha" = "Agri_Value",
 
 # Production 
  "cereal_prod_t"="Production",
  "eggs_prod_t"="Production",
  "fruit_prod_t"="Production",
  "maize_prod_t"="Production",
  "milk_prod_t"="Production",
  "pulses_prod_t"="Production",
  "sugar_crop_prod_t"="Production"
)

pca_data <- merged_df %>%
  select(all_of(vars_pca_per_HA), Continent, ID_Npm_Country) %>%
  drop_na()  # tu peux aussi faire imput si trop restrictif

pca_data <- merged_df %>%
  select(all_of(vars_pca_per_HA), Continent, ID_Npm_Country) %>%
  filter(rowMeans(is.na(.)) <= 0.15)

library( missMDA)

X_incomplete <- scale(pca_data %>% select(all_of(vars_pca_per_HA)))
# Imputation
imputed <- imputePCA(X_incomplete, ncp = 5)  # ou plus de dimensions
X <- imputed$completeObs


```



```{r, echo= FALSE, warning=FALSE, message=FALSE}
res_pca <- prcomp(X, center = TRUE, scale. = TRUE)

```


```{r, echo= FALSE, warning=FALSE, message=FALSE}
fviz_eig(res_pca, addlabels = TRUE, ylim = c(0, 50)) +
  ggtitle("Scree Plot: Variance Explained by Principal Components") +
  theme_minimal()
```

The scree plot below shows the percentage of total variance explained by each principal component. The first two principal components explain approximately 33% and 13% of the variance, respectively, accounting for nearly 46% of the total variability.

While the elbow in the scree plot is not very pronounced (with a slight inflection point after the first component), we proceed by focusing on the first two dimensions to explore the main patterns in the data.

```{r, echo= FALSE, warning=FALSE, message=FALSE}

library(ggrepel)
fviz_pca_var(res_pca,
             col.var = var_categories[colnames(X)],
             #col.var = "contrib",
             palette = "Set2",
             repel = TRUE) +
  theme_minimal()
```

This plot shows how each variable contributes to the two main principal components. The direction and length of each arrow represent the contribution and correlation of the variable with the components.

Variables related to agricultural yields cluster in the bottom-left quadrant, suggesting they contribute strongly to Dimension 2. In contrast, variables associated with energy consumption (coal, petrol, gas) are mostly orthogonal to the yield variables and point toward the upper-left quadrant, indicating a strong contribution to Dimension 1. This suggests that yield and energy consumption are not strongly correlated in this dataset.

Similarly, most input-related variables (insecticides, herbicides) align with energy variables, with the exception of NPK fertilizers, which point in the same direction as the yield variables. This may indicate that NPK use is more closely associated with crop productivity than other inputs.

Interestingly, water consumption variables align with energy usage rather than with yields, suggesting that higher water withdrawal does not necessarily translate into higher agricultural performance at a global scale.

The variable representing the average agricultural production value per hectare (USD PPP) points in an intermediate direction. A reasonable interpretation is that energy consumption is more reflective of general development level (e.g., high energy use in oil-exporting countries with low crop yields), rather than of agricultural performance. In contrast, countries with developed agricultural systems and access to irrigation might achieve both higher yield and higher value-added production per hectare. Hence, the average production value captures both productivity and the diversity of economically valuable crops.


```{r, echo= FALSE, warning=FALSE, message=FALSE}


fviz_pca_ind(res_pca,
             geom = "point",
             col.ind = pca_data$Continent,
             palette = "jco",
             addEllipses = TRUE,
             legend.title = "Continent",
             repel = TRUE) +
  ggtitle("Individuals Factor Map (colored by Continent)") +
  theme_minimal()
```
```{r, include=FALSE}
library(ggrepel)

library(ggplot2)
library(ggrepel)

# Extraire les scores PCA
ind_coords <- as.data.frame(res_pca$x[, 1:2])
ind_coords$Country <- pca_data$ID_Npm_Country

# Plot
ggplot(ind_coords, aes(x = PC1, y = PC2, label = Country)) +
  geom_point(color = "steelblue") +
  geom_text_repel(size = 3, max.overlaps = 100) +
  theme_minimal() +
  ggtitle("Individuals Factor Map (labels instead of legend)")

```

```{r,include=FALSE}
fviz_pca_biplot(res_pca, 
                label = "var", 
                habillage = pca_data$Continent, 
                palette = "jco",
                addEllipses = TRUE, 
                repel = TRUE, 
                col.var = "steelblue")
```









