---
title: "Country Relationships Chord Diagram"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Country Relationships Analysis

This document analyzes relationships between countries based on mentions in country background information.

```{r load-libraries, message=FALSE, warning=FALSE}
# Load required packages
library(readr)     # For reading CSV files
library(circlize)  # For creating chord diagrams
library(RColorBrewer)  # For color palettes
```

```{r load-data}
# Load the CSV file
countries_data <- read_csv("/home/luisa/Documents/University/BHT/Data Visualization/BHTDataVisualizationCourse/data/clean_data.csv", show_col_types = FALSE)

# Convert the CSV data to a more usable format
countries <- list()
for (i in 1:nrow(countries_data)) {
  country_name <- countries_data$Country[i]
  countries[[country_name]] <- as.list(countries_data[i, -1])  # Exclude the country column
}

# Remove "World" from the list of countries if it exists
if ("World" %in% names(countries)) {
  countries <- countries[names(countries) != "World"]
}
```

```{r define-relationships}
# Create a data structure to store country relationships
country_relationships <- list()

# Define alternative names for certain countries
country_alternatives <- list(
  "United Kingdom" = c("Britain", "UK", "England", "British"),
  "United States" = c("USA", "US"),
  "China" = c("PRC", "Chinese", "Chin"),
  "Turkey (Turkiye)" = c("Turkiye", "Turkey", "Turkish"),
  "Congo, Democratic Republic of the" = c("DRC", "DR Congo", "Congo-Kinshasa", "Congo republics"),
  "Congo, Republic of the" = c("Congo-Brazzaville", "Congo republics"),
  "Czechia" = c("Czech Republic", "Czech"),
  "Eswatini" = c("Swaziland", "Swazee"),
  "Burma" = c("Myanmar", "Burmese"),
  "Holy See (Vatican City)" = c("Vatican", "Vatican City"),
  "Micronesia, Federated States of" = c("FSM", "Micronesia"),
  "Gambia, The" = c("The Gambia", "Gambia"),
  "Bahamas, The" = c("The Bahamas", "Bahamas"),
  "Korea, North" = c("North Korea", "DPRK", "N. Korea"),
  "Korea, South" = c("South Korea", "ROK", "S. Korea"),
  "Timor-Leste" = c("East Timor"),
  "Taiwan" = c("Republic of China", "ROC"),
  "Vietnam" = c("Viet Nam"),
  "Germany" = c("German"),
  "France" = c("French"),
  "Italy" = c("Italian"),
  "Spain" = c("Spanish"),
  "Portugal" = c("Portuguese"),
  "Greece" = c("Greek"),
  "Netherlands" = c("Dutch"),
  "Belgium" = c("Belgian"),
  "Sweden" = c("Swedish"),
  "Norway" = c("Norwegian"),
  "Denmark" = c("Danish"),
  "Switzerland" = c("Swiss"),
  "Poland" = c("Polish"),
  "Hungary" = c("Hungarian"),
  "Finland" = c("Finnish"),
  "United Arab Emirates" = c("UAE", "Emirates"),
  "New Zealand" = c("NZ"),
  "Trinidad and Tobago" = c("Trinidad & Tobago")
)
```

```{r find-relationships}
# For each country, find which other countries are mentioned in its background
for (country_name in names(countries)) {
  country_background <- countries[[country_name]]$`Introduction..Background`
  related_countries <- c()
  
  if (!is.null(country_background) && !is.na(country_background)) {
    # Check for mentions of each other country
    for (other_country in names(countries)) {
      # Skip self-references
      if (other_country != country_name) {
        # Check if the other country name appears in the background
        if (grepl(other_country, country_background, ignore.case = TRUE)) {
          related_countries <- c(related_countries, other_country)
        } else {
          # Check alternative names if they exist for this country
          if (other_country %in% names(country_alternatives)) {
            for (alt_name in country_alternatives[[other_country]]) {
              if (grepl(alt_name, country_background, ignore.case = TRUE)) {
                related_countries <- c(related_countries, other_country)
                break  # Found a match, no need to check other alternatives
              }
            }
          }
        }
      }
    }
  }
  
  # Store the relationships for this country
  country_relationships[[country_name]] <- related_countries
}
```

```{r prepare-visualization-data, echo=FALSE}
# Convert the relationships to a matrix format for the chord diagram
countries_list <- unique(names(country_relationships))
n_countries <- length(countries_list)

# Get continent information for each country
country_continents <- sapply(countries_list, function(country) {
  continent <- countries[[country]]$`Geography..Map.references`
  if(is.null(continent) || is.na(continent)) {
    return("Unknown")
  } else {
    return(continent)
  }
})

# Create an empty matrix
relationship_matrix <- matrix(0, nrow = n_countries, ncol = n_countries)
rownames(relationship_matrix) <- countries_list
colnames(relationship_matrix) <- countries_list

# Fill the matrix with relationship data
for (i in 1:n_countries) {
  country_name <- countries_list[i]
  related_countries <- country_relationships[[country_name]]
  
  for (related in related_countries) {
    if (related %in% countries_list) {
      j <- which(countries_list == related)
      relationship_matrix[i, j] <- relationship_matrix[i, j] + 1
    }
  }
}

# Filter to include only countries with at least one relationship
has_relationships <- rowSums(relationship_matrix) > 0 | colSums(relationship_matrix) > 0
filtered_matrix <- relationship_matrix[has_relationships, has_relationships]
filtered_continents <- country_continents[has_relationships]

# Create a more manageable matrix by keeping only countries with multiple relationships
relationship_threshold <- 2
significant_relationships <- rowSums(filtered_matrix) >= relationship_threshold | 
                            colSums(filtered_matrix) >= relationship_threshold
final_matrix <- filtered_matrix[significant_relationships, significant_relationships]
final_continents <- filtered_continents[significant_relationships]

# Sort countries by continent
continent_order <- order(final_continents)
final_matrix <- final_matrix[continent_order, continent_order]
final_continents <- final_continents[continent_order]

# Group countries by continent
unique_continents <- unique(final_continents)
continent_countries <- lapply(unique_continents, function(cont) {
  rownames(final_matrix)[final_continents == cont]
})
names(continent_countries) <- unique_continents

# Create a continent-level matrix
continent_matrix <- matrix(0, nrow = length(unique_continents), ncol = length(unique_continents))
rownames(continent_matrix) <- unique_continents
colnames(continent_matrix) <- unique_continents

# Fill the continent matrix with aggregated relationship data
for (i in 1:nrow(final_matrix)) {
  for (j in 1:ncol(final_matrix)) {
    if (final_matrix[i, j] > 0) {
      source_continent <- final_continents[i]
      target_continent <- final_continents[j]
      continent_matrix[source_continent, target_continent] <- 
        continent_matrix[source_continent, target_continent] + final_matrix[i, j]
    }
  }
}

# Rename "Central America and the Caribbean" to "Central America" in the data
final_continents[final_continents == "Central America and the Caribbean"] <- "Central America"
unique_continents <- unique(final_continents)

# Update continent_countries list with the new name
if ("Central America and the Caribbean" %in% names(continent_countries)) {
  continent_countries[["Central America"]] <- continent_countries[["Central America and the Caribbean"]]
  continent_countries[["Central America and the Caribbean"]] <- NULL
}

# Update continent_matrix with the new name
if ("Central America and the Caribbean" %in% rownames(continent_matrix)) {
  rownames(continent_matrix)[rownames(continent_matrix) == "Central America and the Caribbean"] <- "Central America"
  colnames(continent_matrix)[colnames(continent_matrix) == "Central America and the Caribbean"] <- "Central America"
}
```

## Continent Relationships Chord Diagram

The following chord diagram visualizes the relationships between continents based on country mentions:

```{r chord-diagram, fig.width=12, fig.height=12}
# Create a more pleasant color palette for continents
# Using a custom palette with darker, more saturated colors for better visibility
continent_colors <- c(
  "Africa" = "#0D47A1",       # Darker Blue
  "Asia" = "#B71C1C",         # Darker Red
  "Central America" = "#E65100", # Darker Orange
  "Europe" = "#33691E",       # Darker Green
  "Middle East" = "#D84315",  # Darker Deep Orange
  "North America" = "#4527A0", # Darker Purple
  "Oceania" = "#00695C",      # Darker Teal
  "South America" = "#FF8F00", # Darker Amber
  "Southeast Asia" = "#AD1457", # Darker Pink
  "Arctic Region" = "#006064", # Darker Cyan
  "AsiaEurope" = "#4E342E"    # Darker Brown
)

# Create the chord diagram for continents
par(mar = c(1, 1, 3, 1))
chordDiagram(
  continent_matrix,
  grid.col = continent_colors,
  transparency = 0.1,  
  directional = TRUE,
  direction.type = "arrows",
  link.arr.type = "big.arrow",
  link.sort = TRUE,
  link.largest.ontop = TRUE,
  annotationTrack = "grid",
  preAllocateTracks = list(track.height = 0.1),
  annotationTrackHeight = c(0.05, 0.1),
  grid.border = "black"  # Add thin black borders around sectors
)

# Add labels for continents
circos.track(track.index = 1, panel.fun = function(x, y) {
  sector.index = get.cell.meta.data("sector.index")
  xlim = get.cell.meta.data("xlim")
  ylim = get.cell.meta.data("ylim")
  
  circos.text(
    mean(xlim), ylim[1], 
    sector.index, 
    facing = "clockwise",
    niceFacing = TRUE, 
    adj = c(0, 0.5),
    cex = 0.8
  )
}, bg.border = NA)

# Add title with larger font size
title(main = "Relationships Between Continents", 
      cex.main = 1.8,  # Increase title size
      font.main = 2)   # Bold font
```

## Summary

```{r summary}
# Print a summary of the relationships
cat("Total countries analyzed:", length(countries_list), "\n")
cat("Countries with at least one relationship:", sum(has_relationships), "\n")
cat("Countries in the final visualization:", nrow(final_matrix), "\n")
```

This analysis shows how countries are interconnected based on mentions in their background information. The chord diagram visualizes these relationships at the continent level, highlighting how regions of the world relate to each other. 