---
title: "Global Trade Relations Visualization"
author: "Data Visualization Course"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: united
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

## Introduction

This R Markdown document creates a visualization of global trade relationships, highlighting the top 50 trade relationships while showing others with lower opacity. The visualization uses curved arrows to represent trade flows, with arrow thickness indicating trade volume.

## Loading Required Libraries

```{r libraries}
library(ggplot2)
library(dplyr)
library(readr)
library(stringr)
library(maps)
library(ggrepel)
library(scales)
library(ggarrow)  # For better arrow visualization
library(plotly)   # For interactive plots
```

## Data Loading and Preprocessing

First, we load our dataset containing country information and trade relationships.

```{r load_data}
data <- read_csv("/home/luisa/Documents/University/BHT/Data Visualization/BHTDataVisualizationCourse/data/clean_data.csv", show_col_types = FALSE)
```

## Country Name Standardization

We define alternative names for countries to ensure consistent naming across our dataset.

```{r country_alternatives}
country_alternatives <- list(
  "United Kingdom" = c("Britain", "UK", "England", "British"),
  "United States" = c("USA", "US"),
  "China" = c("PRC", "Chinese", "Chin"),
  "Turkey (Turkiye)" = c("Turkiye", "Turkey", "Turkish"),
  "Congo, Democratic Republic of the" = c("DRC", "DR Congo", "Congo-Kinshasa", "Congo republics"),
  "Congo, Republic of the" = c("Congo-Brazzaville", "Congo republics"),
  "Czechia" = c("Czech Republic", "Czech"),
  "Eswatini" = c("Swaziland", "Swazee"),
  "Burma" = c("Myanmar", "Burmese"),
  "Holy See (Vatican City)" = c("Vatican", "Vatican City"),
  "Micronesia, Federated States of" = c("FSM", "Micronesia"),
  "Gambia, The" = c("The Gambia", "Gambia"),
  "Bahamas, The" = c("The Bahamas", "Bahamas"),
  "Korea, North" = c("North Korea", "DPRK", "N. Korea"),
  "Korea, South" = c("South Korea", "ROK", "S. Korea"),
  "Timor-Leste" = c("East Timor"),
  "Taiwan" = c("Republic of China", "ROC"),
  "Vietnam" = c("Viet Nam"),
  "Germany" = c("German"),
  "France" = c("French"),
  "Italy" = c("Italian"),
  "Spain" = c("Spanish"),
  "Portugal" = c("Portuguese"),
  "Greece" = c("Greek"),
  "Netherlands" = c("Dutch"),
  "Belgium" = c("Belgian"),
  "Sweden" = c("Swedish"),
  "Norway" = c("Norwegian"),
  "Denmark" = c("Danish"),
  "Switzerland" = c("Swiss"),
  "Poland" = c("Polish"),
  "Hungary" = c("Hungarian"),
  "Finland" = c("Finnish"),
  "United Arab Emirates" = c("UAE", "Emirates"),
  "New Zealand" = c("NZ"),
  "Trinidad and Tobago" = c("Trinidad & Tobago")
)
```

## Helper Functions

### Coordinate Parsing
Function to parse coordinates from text format (e.g., "33 00 N, 65 00 E")

```{r parse_coordinates_function}
parse_coordinates <- function(coord_text) {
  if (is.na(coord_text) || coord_text == "") {
    return(list(lat = NA, lon = NA))
  }
  
  # Split by comma
  parts <- str_split(coord_text, ",")[[1]]
  if (length(parts) != 2) {
    return(list(lat = NA, lon = NA))
  }
  
  # Parse latitude
  lat_part <- str_trim(parts[1])
  lat_match <- str_match(lat_part, "([0-9]+)\\s+([0-9]+)\\s+([NS])")
  if (is.na(lat_match[1])) {
    return(list(lat = NA, lon = NA))
  }
  lat_deg <- as.numeric(lat_match[2])
  lat_min <- as.numeric(lat_match[3])
  lat_dir <- lat_match[4]
  lat <- lat_deg + lat_min/60
  if (lat_dir == "S") lat <- -lat
  
  # Parse longitude
  lon_part <- str_trim(parts[2])
  lon_match <- str_match(lon_part, "([0-9]+)\\s+([0-9]+)\\s+([EW])")
  if (is.na(lon_match[1])) {
    return(list(lat = NA, lon = NA))
  }
  lon_deg <- as.numeric(lon_match[2])
  lon_min <- as.numeric(lon_match[3])
  lon_dir <- lon_match[4]
  lon <- lon_deg + lon_min/60
  if (lon_dir == "W") lon <- -lon
  
  return(list(lat = lat, lon = lon))
}
```

### Import Value Extraction
Function to extract numeric values from import data text

```{r extract_import_value_function}
extract_import_value <- function(import_text) {
  if (is.na(import_text) || import_text == "") {
    return(0)
  }
  
  match <- str_match(import_text, "\\$([0-9.]+)\\s*(trillion|billion|million)?")
  if (is.na(match[1])) {
    return(0)
  }
  
  value <- as.numeric(match[2])
  if (!is.na(match[3])) {
    if (match[3] == "million") {
      value <- value / 1000  # Convert to billions
    } else if (match[3] == "trillion") {
      value <- value * 1000  # Convert to billions
    }
  }
  
  return(value)
}
```

### Country Name Standardization
Function to standardize country names using the alternatives list

```{r standardize_country_name_function}
standardize_country_name <- function(country_name) {
  country_name <- str_trim(country_name)
  
  if (country_name %in% names(country_alternatives)) {
    return(country_name)
  }
  
  for (standard_name in names(country_alternatives)) {
    if (country_name %in% country_alternatives[[standard_name]]) {
      return(standard_name)
    }
  }
  
  return(country_name)
}
```

### Trade Partner Parsing
Function to parse trade partners and their percentages from text

```{r parse_trade_partners_function}
parse_trade_partners <- function(partners_text) {
  if (is.na(partners_text) || partners_text == "") {
    return(data.frame())
  }
  
  partners_text <- str_split(partners_text, "note:")[[1]][1]
  partners_text <- str_trim(partners_text)
  
  partners_text <- str_replace(partners_text, "Gambia, The", "Gambia")
  partners_text <- str_replace(partners_text, "Bahamas, The", "Bahamas")
  
  pairs <- str_split(partners_text, ",")[[1]]
  
  result <- data.frame()
  
  for (pair in pairs) {
    percent_match <- str_match(pair, "([0-9.]+)%")
    if (is.na(percent_match[1])) next
    
    percentage <- as.numeric(percent_match[2])
    
    country_part <- str_replace(pair, "\\s*[0-9.]+%.*", "")
    country_name <- str_trim(country_part)
    
    if (country_name != "" && !is.na(percentage)) {
      result <- rbind(result, data.frame(
        partner = standardize_country_name(country_name),
        percentage = percentage
      ))
    }
  }
  
  return(result)
}
```

## Initial Data Visualizations

### Distribution of Import Values
```{r import_distribution, fig.width=10, fig.height=6}
# Create a dataframe with import values
import_distribution <- as.data.frame(data) %>%
  mutate(import_value_billions = sapply(`Economy..Imports`, extract_import_value)) %>%
  filter(!is.na(import_value_billions)) %>%
  # Add small offset to handle zeros and ensure log works
  mutate(import_value_billions = import_value_billions + 0.01) %>%
  arrange(desc(import_value_billions))

# Get top 5 countries for labeling
top_5_countries <- import_distribution %>%
  head(5) %>%
  mutate(
    label = paste0(Country, "\n$", round(import_value_billions - 0.01), "B")  # Subtract offset for display
  )

# Create the violin plot
ggplot(import_distribution, aes(x = "Global Imports", y = import_value_billions)) +
  geom_violin(fill = "#1976D2", alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.4, size = 2, color = "#0D47A1") +
  geom_point(data = top_5_countries, 
             aes(y = import_value_billions),
             color = "red", size = 3) +
  geom_text_repel(
    data = top_5_countries,
    aes(label = label),
    nudge_x = 0.3,
    direction = "y",
    hjust = 0,
    segment.color = "red",
    segment.alpha = 0.5,
    box.padding = 0.5,
    point.padding = 0.3,
    force = 10
  ) +
  labs(
    title = "Distribution of Country Import Values (Log₂ Scale)",
    subtitle = "Top 5 importing countries highlighted in red",
    x = "",
    y = "Import Value (Billions USD) - Log₂ Scale"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold"),
    axis.text = element_text(size = 10),
    axis.title = element_text(size = 12),
    axis.text.x = element_text(angle = 0)
  ) +
  scale_y_continuous(
    trans = scales::log2_trans(),
    labels = function(x) scales::dollar_format(scale = 1, prefix = "$")(x - 0.01),  # Subtract offset for display
    breaks = scales::breaks_log(n = 10, base = 2)
  ) +
  geom_hline(
    yintercept = mean(import_distribution$import_value_billions), 
    color = "#B71C1C", 
    linetype = "dashed", 
    size = 1
  ) +
  annotate(
    "text", 
    x = 0, 
    y = mean(import_distribution$import_value_billions), 
    label = paste0("Mean: $", round(mean(import_distribution$import_value_billions) - 0.01), "B"),  # Subtract offset for display
    color = "#B71C1C",
    vjust = -0.5,
    hjust = -3
  )
```

### Trade Distribution by Region
```{r trade_by_region, fig.width=8, fig.height=8}
# Calculate total trade volume by region
region_trade <- as.data.frame(data) %>%
  mutate(
    import_value_billions = sapply(`Economy..Imports`, extract_import_value),
    region = `Geography..Map.references`
  ) %>%
  group_by(region) %>%
  summarize(
    total_trade = sum(import_value_billions, na.rm = TRUE)
  ) %>%
  arrange(desc(total_trade)) %>%
  filter(!is.na(region)) %>%
  # Calculate percentages for labels
  mutate(
    percentage = round(total_trade / sum(total_trade) * 100, 1),
    hover_text = paste0(
      region, "\n",
      "Trade Volume: $", format(round(total_trade), big.mark = ","), "B\n",
      "Share: ", percentage, "%"
    )
  )

# Create color palette matching the main map
region_colors <- c(
  "North America" = "#0D47A1",
  "Europe" = "#1976D2",
  "Asia" = "#B71C1C",
  "Southeast Asia" = "#E57373",
  "Middle East" = "#E65100",
  "Africa" = "#2E7D32",
  "Central America" = "#FBC02D",
  "South America" = "#F9A825",
  "Oceania" = "#00838F"
)

# Create the interactive pie chart
p <- plot_ly(
  data = region_trade,
  labels = ~region,
  values = ~total_trade,
  type = 'pie',
  textposition = 'inside',
  textinfo = 'label',
  hoverinfo = 'text',
  text = ~hover_text,
  marker = list(colors = region_colors[region_trade$region]),
  hole = 0.4
) %>%
  layout(
    title = list(
      text = "Distribution of Trade Volume by Region (Interactive)",
      font = list(size = 16)
    ),
    showlegend = FALSE,
    annotations = list(
      list(
        text = paste0("Total Global Trade:\n$", format(round(sum(region_trade$total_trade)), big.mark = ","), "B"),
        showarrow = FALSE,
        font = list(size = 12),
        x = 0.5,
        y = 0.5
      )
    )
  )

# Display the interactive plot
p
```

## Data Processing

Process the coordinates and import values from the raw data.

```{r process_data}
print("Processing coordinates...")
coords_list <- lapply(data$`Geography..Geographic.coordinates`, parse_coordinates)
data$lat <- sapply(coords_list, function(x) x$lat)
data$lon <- sapply(coords_list, function(x) x$lon)

print("Processing import values...")
data$import_value_billions <- sapply(data$`Economy..Imports`, extract_import_value)
```

## Creating Trade Relationships Dataset

Process the trade relationships between countries.

```{r create_trade_relationships}
print("Processing trade relationships...")
trade_relationships <- data.frame()

for (i in 1:nrow(data)) {
  country <- data$Country[i]
  import_value <- data$import_value_billions[i]
  country_lat <- data$lat[i]
  country_lon <- data$lon[i]
  
  if (is.na(country_lat) || is.na(country_lon) || import_value == 0) {
    next
  }
  
  partners <- parse_trade_partners(data$`Economy..Imports...partners`[i])
  
  if (nrow(partners) > 0) {
    for (j in 1:nrow(partners)) {
      partner_name <- partners$partner[j]
      percentage <- partners$percentage[j]
      
      partner_row <- which(data$Country == partner_name)
      if (length(partner_row) == 0) {
        found <- FALSE
        for (alt_name in names(country_alternatives)) {
          if (partner_name %in% country_alternatives[[alt_name]]) {
            partner_row <- which(data$Country == alt_name)
            if (length(partner_row) > 0) {
              found <- TRUE
              break
            }
          }
        }
        if (!found) {
          # cat("Country not found:", partner_name, "\n")
          next
        }
      }
      
      partner_lat <- data$lat[partner_row[1]]
      partner_lon <- data$lon[partner_row[1]]
      
      if (!is.na(partner_lat) && !is.na(partner_lon)) {
        line_thickness <- (percentage / 100) * import_value
        distance <- sqrt((partner_lon - country_lon)^2 + (partner_lat - country_lat)^2)
        
        if (distance < 10 && line_thickness > 20) {
          curvature_category <- "extreme"
          curvature_value <- 1.2
        } else if (distance < 10 || line_thickness > 10) {
          curvature_category <- "very_high"
          curvature_value <- 0.9
        } else if (distance < 10) {
          curvature_category <- "high"
          curvature_value <- 0.6
        } else if (distance < 30 || line_thickness > 5) {
          curvature_category <- "medium_high"
          curvature_value <- 0.9
        } else if (distance < 30) {
          curvature_category <- "medium"
          curvature_value <- 0.3
        } else {
          curvature_category <- "low" 
          curvature_value <- 0.1
        }
        
        trade_relationships <- rbind(trade_relationships, data.frame(
          from_country = country,
          to_country = partner_name,
          from_lat = country_lat,
          from_lon = country_lon,
          to_lat = partner_lat,
          to_lon = partner_lon,
          percentage = percentage,
          import_value = import_value,
          line_thickness = line_thickness,
          curvature_category = curvature_category,
          curvature = curvature_value,
          arrow_size = 0.2 + (0.1 * log1p(line_thickness)),
          distance = distance
        ))
      }
    }
  }
}

print(paste("Created", nrow(trade_relationships), "trade relationships"))
```

## Identifying Top Trade Relationships

Sort and mark the top 50 trade relationships.

```{r identify_top_50}
trade_relationships <- trade_relationships[order(-trade_relationships$line_thickness), ]
trade_relationships$is_top_50 <- 1:nrow(trade_relationships) <= 50

print(paste("Top 50 relationships range from", 
            round(min(trade_relationships$line_thickness[trade_relationships$is_top_50]), 2),
            "to", 
            round(max(trade_relationships$line_thickness[trade_relationships$is_top_50]), 2),
            "billion USD"))
```

## Preparing Arrow Path Data

Helper functions for creating curved arrows.

```{r arrow_helper_functions}
get_bend_factor <- function(level, distance) {
  base_levels <- list(
    "low" = 3.5,         
    "medium" = 3.5,      
    "medium_high" = 4, 
    "high" = 4.5,        
    "very_high" = 5,   
    "extreme" = 5.5      
  )
  distance_factor <- 1 + log1p(distance/20)
  return(base_levels[[level]] * distance_factor)
}

get_bend_direction <- function(start_lon, start_lat, end_lon, end_lat) {
  if (abs(start_lon - end_lon) < 5) {
    if (end_lat > start_lat) return(1) else return(-1)
  }
  if (end_lon > start_lon) return(1) else return(-1)
}
```

Create the arrow path data.

```{r create_arrow_paths}
print("Creating arrow paths...")
arrow_data <- data.frame()
t <- seq(0, 1, length.out = 50)

for (i in 1:nrow(trade_relationships)) {
  rel <- trade_relationships[i,]
  
  from_lon <- rel$to_lon
  to_lon <- rel$from_lon
  from_lat <- rel$to_lat
  to_lat <- rel$from_lat
  
  x_range <- to_lon - from_lon
  y_range <- to_lat - from_lat
  
  distance <- sqrt(x_range^2 + y_range^2)
  
  bend_dir <- get_bend_direction(from_lon, from_lat, to_lon, to_lat)
  bend_factor <- get_bend_factor(rel$curvature_category, distance)
  
  is_vertical <- abs(x_range) < 5
  if (is_vertical) {
    x <- from_lon + x_range * t + bend_dir * bend_factor * sin(pi * t) * 1.2
    y <- from_lat + y_range * t
  } else {
    x <- from_lon + x_range * t
    y <- from_lat + y_range * t + bend_dir * bend_factor * sin(pi * t)
  }
  
  width_scaled <- sqrt(rel$line_thickness) * 2
  alpha_value <- ifelse(rel$is_top_50, 0.8, 0.3)
  
  arrow_data <- rbind(arrow_data, data.frame(
    x = x,
    y = y,
    group = i,
    width = width_scaled,
    is_top_50 = rel$is_top_50,
    alpha_value = alpha_value
  ))
}
```

## Creating the Visualization

Create and save the final visualization.

```{r create_visualization, fig.width=20, fig.height=12, out.width='100%'}
# Create a mapping of map references to continents and colors
continent_colors <- c(
  # Western-aligned regions in blue spectrum
  "North America" = "#0D47A1",    # Dark blue
  "Europe" = "#1976D2",           # Medium blue
  
  # Asian regions in red spectrum
  "Asia" = "#B71C1C",             # Dark red
  "Southeast Asia" = "#E57373",   # Lighter red
  
  # Updated colors as requested
  "Middle East" = "#E65100",      # Orange
  "Africa" = "#2E7D32",           # Green
  "Central America" = "#FBC02D",  # Yellow
  "South America" = "#F9A825",    # Darker yellow
  
  # Other regions
  "Oceania" = "#00838F",          # Teal
  "Unknown" = "#757575"           # Gray for unknown/missing values
)

# Handle missing values and create node sizes directly
valid_data <- !is.na(data$lat) & !is.na(data$lon)
node_sizes <- rep(1, nrow(data))  # Default size for all points
trade_volumes <- data$import_value_billions
trade_volumes[is.na(trade_volumes)] <- 0
max_volume <- max(trade_volumes, na.rm = TRUE)
node_sizes[valid_data] <- 1 + 4 * (trade_volumes[valid_data] / max_volume)

# Handle missing continent data
data$continent <- data$`Geography..Map.references`
data$continent[is.na(data$continent)] <- "Unknown"

world_map <- map_data("world")

p <- ggplot() +
  geom_polygon(data = world_map, 
               aes(x = long, y = lat, group = group), 
               fill = "lightgray", 
               color = "white", 
               size = 0.2) +
  
  geom_arrow(data = arrow_data,
             aes(x = x, y = y, 
                 group = group,
                 linewidth = width * 1.5,
                 alpha = alpha_value * width/50),  
             color = "cyan4") +
  
  geom_point(data = data[valid_data, ],
             aes(x = lon, y = lat,
                 color = continent,
                 size = node_sizes[valid_data]),
             alpha = 0.8) +
  
  scale_color_manual(name = "Continent/Region",
                    values = continent_colors,
                    guide = guide_legend(override.aes = list(size = 8))) +
  
  scale_size_continuous(name = "Trade Volume",
                       range = c(1, 5),
                       breaks = c(1, 3, 5),
                       labels = c("No data/0", 
                                "$500B",
                                "$5T"),
                       guide = guide_legend(override.aes = list(size = c(0.6, 3, 5)))) +
  
  scale_linewidth_continuous(name = "Trade Flow Volume\n(Billions USD)",
                         range = c(0.2, 3),
                         breaks = c(10, 50),
                         labels = c("$50B", "$500B"),
                         guide = guide_legend(override.aes = list(alpha = 1))) +
  
  scale_alpha_identity() +
  
  theme_void() +
  theme(
    plot.title = element_text(size = 26, hjust = 0.5, margin = margin(b = 20)),
    plot.subtitle = element_text(size = 18, hjust = 0.5, margin = margin(b = 20)),
    legend.position = "right",
    legend.box = "vertical",
    legend.margin = margin(10, 10, 10, 10),
    legend.box.spacing = unit(2, "lines"),
    legend.text = element_text(size = 16),
    legend.title = element_text(size = 18, face = "bold"),
    legend.key.size = unit(2.5, "lines"),
    legend.key.height = unit(2.5, "lines"),
    legend.spacing.y = unit(2, "lines"),
    panel.background = element_rect(fill = "lightblue", color = NA)
  ) +
  
  labs(
    title = "Global Trade Relationships",
    subtitle = "Directed import flows: arrows point from trade partners to importing countries\nTop 50 relationships by trade volume highlighted, others shown with low opacity",
    caption = "Arrow thickness = Import Volume (billions) × Trade Share (%)\nArrows point from trade partner to importing country\nNode size indicates total trade volume (smallest nodes represent countries with no data or zero trade)"
  ) +
  
  coord_fixed(ratio = 1.3, xlim = c(-180, 180), ylim = c(-60, 80))

```

## Display the plot

```{r display_plot, fig.width=20, fig.height=12, out.width='100%'}
# Display the plot
ggsave("global_trade_relationships.png", p, width = 20, height = 12, dpi = 300)
print(p)
```

