---
title: "PaperReadyQuestion4"
author: "Nick Chandler"
date: "2025-07-09"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(dplyr)
library(ggplot2)
library(ggrepel)
library(tidyr)
library(ks)
library(rgl)
library(scales)
library(MASS)
library(plotly)
library(fmsb)
library(patchwork)
library(tidyverse)
library(aplpack)
library(leaflet)
library(rnaturalearth)
library(countrycode)
library(cowplot)
```


```{r}
col_names_and_nas <- function(df) {
  for (i in seq_along(df)) {
    na_count <- sum(is.na(df[[i]]))
    empty_count <- sum(df[[i]] == "", na.rm = TRUE)
    cat(colnames(df)[i], ":", na_count + empty_count, "\n")
  }
}

# Read in data
full_df <- read.csv("../data/clean_data.csv", header = TRUE)

# Trim data
sub_cols <- grep("male|female", colnames(full_df),ignore.case=TRUE)
cols_to_keep <- colnames(full_df[, sub_cols])
cols_to_keep <- c(cols_to_keep, c("Geography..Map.references", "Geography..Geographic.coordinates", "Economy..Real.GDP.per.capita", "People.and.Society..Population...total", "People.and.Society..Birth.rate", "People.and.Society..Maternal.mortality.ratio", "People.and.Society..Infant.mortality.rate...total", "People.and.Society..Life.expectancy.at.birth...total.population", "People.and.Society..Population.growth.rate"))

trim_df <- full_df[-78,cols_to_keep]  # Note, we omit vatican city here due to lack of gender specific data
trim_df$Country <- full_df[-78,1]
rownames(trim_df) <- full_df[-78,1]

# Status report
#col_names_and_nas(trim_df)
```

```{r}
# Economy..Real.GDP.per.capita
trim_df$Economy..Real.GDP.per.capita <- as.numeric(gsub("[$,]", "", str_extract(trim_df$Economy..Real.GDP.per.capita, "\\$[0-9,]+")))
colnames(trim_df)[colnames(trim_df) == "Economy..Real.GDP.per.capita"] <- "Real.GDP.per.capita.USD"

# People.and.Society..Population.growth.rate
trim_df$People.and.Society..Population.growth.rate <- as.numeric(sub("%", "", str_extract(trim_df$People.and.Society..Population.growth.rate, "[0-9.]+%"))) / 100
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Population.growth.rate"] <- "Population.growth.rate.percent"

# People.and.Society..Birth.rate - births 
trim_df$People.and.Society..Birth.rate <- as.numeric(str_extract(trim_df$People.and.Society..Birth.rate, "^[0-9]+")) / 1000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Birth.rate"] <- "Live.births.per.1000.people"

# People.and.Society..Population...total
trim_df$People.and.Society..Population...total <- as.numeric(gsub(",", "", str_extract(trim_df$People.and.Society..Population...total, "^[0-9,]+")))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Population...total"] <- "Total.population"

# People.and.Society..Maternal.mortality.ratio - deaths per live birth
trim_df$People.and.Society..Maternal.mortality.ratio <- as.numeric(gsub(",", "", str_extract(trim_df$People.and.Society..Maternal.mortality.ratio, "^[0-9]+"))) / 100000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Maternal.mortality.ratio"] <- "Maternal.deaths.per.100000.live.births"

# People.and.Society..Infant.mortality.rate...total - deaths per live birth
trim_df$People.and.Society..Infant.mortality.rate...total <- as.numeric(str_extract(trim_df$People.and.Society..Infant.mortality.rate...total, "^[0-9.]+")) / 1000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Infant.mortality.rate...total"] <- "Infant.deaths.per.1000.live.births"

# People.and.society..Life.expectancy.at.birth...total.population
trim_df$People.and.Society..Life.expectancy.at.birth...total.population <- as.numeric(str_extract(trim_df$People.and.Society..Life.expectancy.at.birth...total.population, "^[0-9.]+"))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Life.expectancy.at.birth...total.population"] <- "Life.expectancy.at.birth"

# Geography..Map.references
colnames(trim_df)[colnames(trim_df) == "Geography..Map.references"] <- "Continent"

# People.and.Society..Population...male
trim_df$People.and.Society..Population...male <- as.numeric(gsub(",", "", str_extract(trim_df$People.and.Society..Population...male, "^\\s*[0-9,]+")))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Population...male"] <- "Male.population"

# People.and.Society..Population...female
trim_df$People.and.Society..Population...female <- as.numeric(gsub(",", "", str_extract(trim_df$People.and.Society..Population...female, "^\\s*[0-9,]+")))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Population...female"] <- "Female.population"

# Economy..Youth.unemployment.rate..ages.15.24....female
trim_df$Economy..Youth.unemployment.rate..ages.15.24....female <- as.numeric(str_extract(trim_df$Economy..Youth.unemployment.rate..ages.15.24....female, "^\\s*[0-9.]+")) / 100
colnames(trim_df)[colnames(trim_df) == "Economy..Youth.unemployment.rate..ages.15.24....female"] <- "Female.unemployment.rate.15.24"

# Economy..Youth.unemployment.rate..ages.15.24....male
trim_df$Economy..Youth.unemployment.rate..ages.15.24....male <- as.numeric(str_extract(trim_df$Economy..Youth.unemployment.rate..ages.15.24....male, "^\\s*[0-9.]+")) / 100
colnames(trim_df)[colnames(trim_df) == "Economy..Youth.unemployment.rate..ages.15.24....male"] <- "Male.unemployment.rate.15.24"


# People.and.Society..School.life.expectancy..primary.to.tertiary.education....female
trim_df$People.and.Society..School.life.expectancy..primary.to.tertiary.education....female <- as.numeric(str_extract(trim_df$People.and.Society..School.life.expectancy..primary.to.tertiary.education....female, "^\\s*[0-9]+"))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..School.life.expectancy..primary.to.tertiary.education....female"] <- "Female.expected.years.of.education"

# People.and.Society..School.life.expectancy..primary.to.tertiary.education....male
trim_df$People.and.Society..School.life.expectancy..primary.to.tertiary.education....male <- as.numeric(str_extract(trim_df$People.and.Society..School.life.expectancy..primary.to.tertiary.education....male, "^\\s*[0-9]+"))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..School.life.expectancy..primary.to.tertiary.education....male"] <- "Male.expected.years.of.education"

# People.and.Society..Literacy...female
trim_df$People.and.Society..Literacy...female <- as.numeric(str_extract(trim_df$People.and.Society..Literacy...female, "^\\s*[0-9.]+")) / 100
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Literacy...female"] <- "Female.literacy.rate"

# People.and.Society..Literacy...male
trim_df$People.and.Society..Literacy...male <- as.numeric(str_extract(trim_df$People.and.Society..Literacy...male, "^\\s*[0-9.]+")) / 100
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Literacy...male"] <- "Male.literacy.rate"

# People.and.Society..Tobacco.use...female
trim_df$People.and.Society..Tobacco.use...female <- as.numeric(str_extract(trim_df$People.and.Society..Tobacco.use...female, "^\\s*[0-9.]+")) / 100
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Tobacco.use...female"] <- "Female.proportion.of.tobacco.users"

# People.and.Society..Tobacco.use...male
trim_df$People.and.Society..Tobacco.use...male <- as.numeric(str_extract(trim_df$People.and.Society..Tobacco.use...male, "^\\s*[0-9.]+")) / 100
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Tobacco.use...male"] <- "Male.proportion.of.tobacco.users"

# People.and.Society..Life.expectancy.at.birth...female
trim_df$People.and.Society..Life.expectancy.at.birth...female <- as.numeric(str_extract(trim_df$People.and.Society..Life.expectancy.at.birth...female, "^\\s*[0-9.]+"))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Life.expectancy.at.birth...female"] <- "Female.life.expectancy.at.birth"

# People.and.Society..Life.expectancy.at.birth...male
trim_df$People.and.Society..Life.expectancy.at.birth...male <- as.numeric(str_extract(trim_df$People.and.Society..Life.expectancy.at.birth...male, "^\\s*[0-9.]+"))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Life.expectancy.at.birth...male"] <- "Male.life.expectancy.at.birth"

# People.and.Society..Infant.mortality.rate...female
trim_df$People.and.Society..Infant.mortality.rate...female <- as.numeric( str_extract(trim_df$People.and.Society..Infant.mortality.rate...female, "^\\s*[0-9.]+")) / 1000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Infant.mortality.rate...female"] <- "Female.IMR.deaths.per.birth"

# People.and.Society..Infant.mortality.rate...male
trim_df$People.and.Society..Infant.mortality.rate...male <- as.numeric( str_extract(trim_df$People.and.Society..Infant.mortality.rate...male, "^\\s*[0-9.]+")) / 1000
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Infant.mortality.rate...male"] <- "Male.IMR.deaths.per.birth"

# People.and.Society..Median.age...female
trim_df$People.and.Society..Median.age...female <- as.numeric( str_extract(trim_df$People.and.Society..Median.age...female, "^\\s*[0-9.]+"))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Median.age...female"] <- "Median.age.female"

# People.and.Society..Median.age...male
trim_df$People.and.Society..Median.age...male <- as.numeric( str_extract(trim_df$People.and.Society..Median.age...male, "^\\s*[0-9.]+"))
colnames(trim_df)[colnames(trim_df) == "People.and.Society..Median.age...male"] <- "Median.age.male"
```


# Introduction

In this section we seek to answer the question: "How do selected demographic variables differ between males and females?". The variables we examine are: Population, Life Expectancy at Birth, Infant Mortality Rate, and Median Age. The techniques we use are Pairs Plots, Kernel Density Estimation + Rug Plots, PCA, and Choropleths.

# Correlation

```{r, fig.height=8, fig.width=10}
gendered_df <- trim_df[,c("Male.population", "Female.population", "Median.age.male", "Median.age.female", "Male.IMR.deaths.per.birth", "Female.IMR.deaths.per.birth", "Male.life.expectancy.at.birth", "Female.life.expectancy.at.birth")]

par(mar = c(8,8,3,3))  # Modify margins to make the plot larger (top, left, bottom, right)
colnames(gendered_df)
colnames(gendered_df) <- c("Male.population", "Female.population", "Male.median.age", "Female.median.age", "Male IMR", "Female IMR", "M.Life.Expectancy", "F.Life.Expectancy")
colnames(gendered_df)

pairs(gendered_df,
      labels = colnames(gendered_df),
      cex.labels = 0.8,            # label size
      font.labels = 2,             # bold
      gap = 1,                   # spacing between plots and labels
      upper.panel = NULL,
      pch = 21, 
      bg = "lightblue")         # optional: simplify upper triangle
```
In these plots, one can see a strong, positive linear relationship between the male and female values for each of the variables: population, median age, infant mortality rate, and life expectancy per country. Because of this, we focus on these variables for the remainder of the analysis.


# Univariate Statistical Graphics

```{r}
plot_two_kde_plots_on_same_axes <- function(x1, x2,
                                            label1,
                                            label2,
                                            title, 
                                            xlab,
                                            col1 = "pink",
                                            col2 = "lightblue", 
                                            lwd=2, legend.pos="topright", rug.alpha=0.4, ...) {
  x1 <- na.omit(x1)
  x2 <- na.omit(x2)
  
  xlim = range(c(x1,x2))
  
  plot(density(x1),
       xlab=xlab, main=title, xlim=xlim,
       col=col1, lwd=lwd, ...)
  lines(density(x2), col=col2, lwd=lwd)
  legend(legend.pos, legend=c(label1, label2), col=c(col1, col2), lwd=lwd)
  
  # Add a color coded rug
  rug(x1, col=adjustcolor(col1, alpha.f=rug.alpha), side=1)
  rug(x2, col=adjustcolor(col2, alpha.f=rug.alpha), side=1)
}
```

### Median Age Across Genders
```{r, fig.width=10, fig.height=10}
par(mfrow=c(2,2))

# Median Age
plot_two_kde_plots_on_same_axes(trim_df$Median.age.female, trim_df$Median.age.male, "Female", "Male", "Median Age of Females and Males", "Median Age (years since birth)")

# Infant Mortality Rate
plot_two_kde_plots_on_same_axes(trim_df$Female.IMR.deaths.per.birth, trim_df$Male.IMR.deaths.per.birth, "Female", "Male", "Infant Mortality Ratio (IMR) of Males and Females", "IMR (Deaths per Birth)")

# Life Expectancy
plot_two_kde_plots_on_same_axes(trim_df$Female.life.expectancy.at.birth, trim_df$Male.life.expectancy.at.birth, "Female", "Male", "Life Expectancy of Males and Females", "Expected Years of Life After Birth", legend.pos="topleft")

# Population
plot_two_kde_plots_on_same_axes(trim_df$Female.population, trim_df$Male.population, "Female", "Male", "Population of Males and Females", "Population")
```


```{r}

plot_gender_by_continent <- function(data, male_var, female_var, title, y_label) {
  library(dplyr)
  library(tidyr)
  library(ggplot2)
  
  agg_df <- data %>%
    group_by(Continent) %>%
    summarise(
      Male = mean(.data[[male_var]], na.rm = TRUE),
      Female = mean(.data[[female_var]], na.rm = TRUE)
    ) %>%
    mutate(Total = Male + Female) %>%
    arrange(desc(Total)) %>%
    select(-Total) %>%
    pivot_longer(cols = c(Male, Female), names_to = "Gender", values_to = "Value")
  
  ggplot(agg_df, aes(x = reorder(Continent, -Value, sum), y = Value, fill = Gender)) +
    geom_bar(stat = "identity", position = "dodge") +
    labs(
      title = title,
      x = "Continent",
      y = y_label
    ) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    scale_fill_manual(values = c("Male" = "lightblue", "Female" = "pink"))
}
```

```{r, fig.width=10, fig.height=10}
age <- plot_gender_by_continent(trim_df, "Median.age.male", "Median.age.female", title="Median Age per Continent", y_label="Average Age (years)")

imr <- plot_gender_by_continent(trim_df, "Male.IMR.deaths.per.birth", "Female.IMR.deaths.per.birth", title="Infant Mortality Rate per Continent", y_label="Average IMR (deaths per birth)")

life.expectancy <- plot_gender_by_continent(trim_df, "Male.life.expectancy.at.birth", "Female.life.expectancy.at.birth", title="Life Expectancy per Continent", y_label="Average Life Expectancy (years)")

population <- plot_gender_by_continent(trim_df, "Male.population", "Female.population", title="Population per Continent", y_label="Average Population")

plot_grid(age, imr, life.expectancy, population, ncol = 2)
```




# Choropleths

Here we present choropleths that display the absolute difference between the variables for each country.

```{r}
trim_df <- trim_df %>%
  mutate(
    PopDiff = abs(Male.population - Female.population),
    IMRDiff = abs(Male.IMR.deaths.per.birth - Female.IMR.deaths.per.birth),
    MedianAgeDiff = abs(Median.age.male - Median.age.female),
    LifeExpectancyDiff = abs(Male.life.expectancy.at.birth - Female.life.expectancy.at.birth)
  )

plot_static_country_data <- function(df, country_col, value_col, 
                                     fill_palette = "YlOrRd", 
                                     legend_title = NULL, 
                                     missing_color = "gray90") {
  library(sf)
  library(rnaturalearth)
  library(rnaturalearthdata)
  library(countrycode)
  library(dplyr)
  library(ggplot2)
  
  # Step 1: Clean + add ISO3 codes
  df <- na.omit(df)
  df$iso_a3 <- countrycode(df[[country_col]], "country.name", "iso3c")
  df$iso_a3[df[[country_col]] == "Kosovo"] <- "XKX"

  # Step 2: Load map
  world <- ne_countries(returnclass = "sf", scale = "medium")
  
  # Step 3: Merge
  joined <- left_join(world, df[, c("iso_a3", value_col)], by = "iso_a3")

  # Step 4: Plot
  ggplot(joined) +
    geom_sf(aes(fill = .data[[value_col]]), color = "white", size = 0.1) +
    scale_fill_distiller(palette = fill_palette, direction = 1, na.value = missing_color) +
    labs(fill = legend_title %||% value_col) +
    theme_void() +
    theme(legend.position = "right")
}
```


```{r, fig.width=10, fig.height=10}
ma.map <- plot_static_country_data(
  df = trim_df,
  country_col = "Country",
  value_col = "MedianAgeDiff",
  fill_palette = "RdPu",
  legend_title = "Abs. Difference in Median Ages"
)

imr.map <- plot_static_country_data(
  df = trim_df,
  country_col = "Country",
  value_col = "IMRDiff",
  fill_palette = "RdPu",
  legend_title = "Abs. Difference in Infant Mortality Rate"
)

le.map <- plot_static_country_data(
  df = trim_df,
  country_col = "Country",
  value_col = "LifeExpectancyDiff",
  fill_palette = "RdPu",
  legend_title = "Abs. Difference in Life Expectancy"
)

pop.map <- plot_static_country_data(
  df = trim_df,
  country_col = "Country",
  value_col = "PopDiff",
  fill_palette = "RdPu",
  legend_title = "Abs. Difference in Population"
)


(ma.map | imr.map) / (le.map | pop.map)
```


# Conclusion


